---
title: "ASESII 24A02 Final Project"
subtitle: "Predict noise level generated by NACA airfoil blade sections"
author: |
  | Group 02
  | Nguyen Thi Yen Nhi (24125071)
  | Nguyen Khang Thinh (24125104)
  | Nguyen Van Tinh (24125106)
  | Nguyen Le Quoc Huy (24125100)
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
link-citations: true
params:
  group_number: 02
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere}
  - \usepackage{float}
  - \usepackage{longtable}
---




# Authorship and contribution statement {-}

<!-- We confirm that every group member can explain the submitted analysis. AI use is reported in `Group04_AI_Log.csv`, and the group checked every AI-assisted item used in this report. -->

```{=latex}
  \begingroup
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{p{0.25\linewidth} p{0.09\linewidth} p{0.57\linewidth}}
  \hline
  Student & ID & Main contributions  \\
  \hline
  Nguyen Thi Yen Nhi & 24125071 &
  Target and predictors, split and preprocessing plan, and data exploration  \\
  \hline
  Nguyen Khang Thinh & 24125104 &
  Dataset selection, research questions, hypotheses, and descriptive statistics for the experimental design \\
  \hline
  Nguyen Van Tinh & 24125106 &
  OLS, Ridge, and Lasso regression, including model fitting, regularization, coefficient analysis, and model comparison \\
  \hline
  Nguyen Le Quoc Huy & 24125100 &
  ANOVA/factorial design analysis, post-hoc analysis, and conclusion and evaluation \\
  \hline
  \end{tabular}
  \endgroup
  ```

# Abstract {-}

This study investigates the prediction of sound pressure levels generated by NACA 0012 airfoil blade sections using five aerodynamic and geometric predictors. The dataset contains 1,503 wind-tunnel observations with no missing values, duplicate rows, or physically invalid measurements. Exploratory analysis revealed weak individual predictor-response relationships and moderate correlation between attack angle and suction-side displacement thickness, while the frequency-response relationship showed evidence of nonlinearity.

We compared Ordinary Least Squares (OLS), Ridge, and Lasso regression using an 80/20 train-test split and five-fold cross-validation. Predictors were standardized using training-set statistics only to prevent data leakage. OLS achieved a cross-validation MSE of 23.03, while Ridge and Lasso produced similar predictive errors. However, Ridge substantially improved numerical conditioning, reducing the condition number from 12.30 for OLS to 5.34 at its optimal penalty. Lasso did not eliminate any of the five original predictors at either tuning rule.

A perturbation experiment introduced a near-duplicate predictor highly correlated with frequency (r=0.995) to evaluate model stability under stronger multicollinearity. Ridge maintained stable predictive performance by redistributing coefficients across correlated predictors, whereas Lasso demonstrated variable-selection behavior under the more conservative 1SE rule.

Overall, the results show that regularization provides greater numerical stability without sacrificing substantial predictive accuracy, with Ridge offering the most consistent response to multicollinearity in this dataset.


```{r setup, echo=FALSE, warning=FALSE, message = FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr", "car", "ggplot2", "dplyr", "kableExtra", "emmeans")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)
library(car)
library(ggplot2)
library(kableExtra)
library(dplyr)
library(emmeans)
# `params` exists only while knitting; the fallback
# keeps interactive chunk runs working.
has_params <- exists("params") && !is.null(params$group_number)
group_number <- if (has_params) as.integer(params$group_number) else 4L
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```

```{r helpers, echo=FALSE}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
analysis_model <- function(x, y, alpha, foldid) {
  model <- fit_cv_glmnet(x = x, y = y, alpha=alpha, foldid = foldid)
  list(
    model = model,
    lambda_min = model$lambda.min,
    mse_min = model$cvm[model$lambda == model$lambda.min],
    cond_min = safe_condition_numbers(x, lambda = model$lambda.min)[2],
    coef_min = coef(model, "lambda.min"),
    coef_min_norm = sqrt(sum(coef(model, "lambda.min")[-1]^2)),
    nonzero_min = count_nonzero(model, s="lambda.min"),
    lambda_1se = model$lambda.1se,
    cond_1se = safe_condition_numbers(x, lambda = model$lambda.1se)[2],
    mse_1se = model$cvm[model$lambda == model$lambda.1se],
    coef_1se = coef(model, "lambda.1se"),
    coef_1se_norm = sqrt(sum(coef(model, "lambda.1se")[-1]^2)),
    nonzero_1se = count_nonzero(model, s="lambda.1se")
  )
}
```

# Prediction design
## Target and predictors
```{r data-configuration, echo=FALSE}

data <- read.table("data/airfoil_self_noise.dat", header = FALSE)
colnames(data) <- c("frequency", "attack_angle", "chord_length", "free_stream_velocity", "suction_side_displacement_thickness", "scaled_sound_pressure")

write.csv(
  data,
  "data/airfoil_self_noise.csv",
  row.names = FALSE
)

config <- list(
  file = "airfoil_self_noise.csv",
  response = "scaled_sound_pressure",
  excluded = "scaled_sound_pressure"
)
data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```


```{r data-overview, echo=FALSE, warning=FALSE, message=FALSE}

data_overview <- tibble(
  Variable = names(data_raw),
  Role     = ifelse(names(data_raw) == config$response, "Response", "Predictor"),
  Type     = vapply(data_raw, function(x) class(x)[1L], character(1)),
  Distinct = vapply(data_raw, dplyr::n_distinct, integer(1), na.rm = TRUE),
  Missing  = vapply(data_raw, function(x) sum(is.na(x)), integer(1))
)

kable(
  data_overview,
  caption = paste("Response and candidate predictors, with the integrity checks",
                  "run before the split.")
)

n_duplicate_rows <- sum(duplicated(analysis_data))
n_missing_total  <- sum(is.na(analysis_data))
n_levels_freq    <- dplyr::n_distinct(data_raw$frequency)
n_levels_chord   <- dplyr::n_distinct(data_raw$chord_length)
n_levels_vel     <- dplyr::n_distinct(data_raw$free_stream_velocity)
n_levels_angle   <- dplyr::n_distinct(data_raw$attack_angle)
n_levels_thick   <- dplyr::n_distinct(data_raw$suction_side_displacement_thickness)

kable(
  data.frame(
    Check = c("Missing values (total cells)", "Fully duplicated rows"),
    Count = c(n_missing_total, n_duplicate_rows)
  ),
  caption = "Missing-value and duplicate-row checks on the raw data."
)
```

The dataset contains `r nrow(data_raw)` observations of `r ncol(data_raw)` numeric
variables: five predictors recorded during NACA 0012 airfoil wind-tunnel tests
and the response `scaled_sound_pressure`, the scaled sound pressure level in
decibels. No missing values or duplicated rows are present (table above), so no
imputation or de-duplication is required before the split.

## Split and preprocessing plan
```{r split-and-scale, echo=FALSE}
split <- split_rows(nrow(analysis_data), seed = seeds$split)
train_id <- split$train
test_id <- split$test
train_df <- analysis_data[train_id, ]

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]

foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

To ensure a fair comparison among OLS, ridge, lasso, and elastic net, all preprocessing and model development steps were performed using only the training data. The holdout test set was reserved exclusively for the final evaluation and was not used during model fitting, model selection, or hyperparameter tuning.
```{r echo=FALSE, warning=FALSE, message=FALSE}
obs <- c(20,35,20,25); probs <- rep(1/4,4)
chisq.test(x = obs, p = probs)
```

### Missing-value handling
```{r echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE}
kable(data.frame(Total_missing_values = sum(is.na(analysis_data))))
```
No missing values were detected; therefore, no imputation was required. If missing values had been present, imputation statistics would have been estimated using only the training data and then applied unchanged to the test data.
### Invalid-value check

```{r invalid-values, echo=FALSE, warning=FALSE, message=FALSE}
implausible_checks <- data.frame(
  Variable   = c("frequency", "attack_angle", "chord_length",
                 "free_stream_velocity", "suction_side_displacement_thickness"),
  Rule       = c("> 0 Hz", ">= 0 deg", "> 0 m", "> 0 m/s", "> 0 m"),
  Violations = c(
    sum(train_df$frequency <= 0),
    sum(train_df$attack_angle < 0),
    sum(train_df$chord_length <= 0),
    sum(train_df$free_stream_velocity <= 0),
    sum(train_df$suction_side_displacement_thickness <= 0)
  )
)

kable(
  implausible_checks,
  caption = "Physical-plausibility checks on the training predictors."
)
```

Beyond the statistical IQR screen below, each predictor was also checked against
the range it is physically allowed to take: a frequency, chord length, velocity,
or boundary-layer thickness cannot be zero or negative, and an angle of attack
cannot be negative. No violations are found on the training set, which is
expected for an instrument-recorded wind-tunnel experiment rather than
manually-entered data.

### Outlier detection and handling

```{r outlier-detection, echo=FALSE, warning=FALSE, message=FALSE}
detect_outliers_iqr <- function(x) {
  q     <- stats::quantile(x, probs = c(0.25, 0.75), na.rm = TRUE)
  fence <- 1.5 * (q[2] - q[1])
  which(x < q[1] - fence | x > q[2] + fence)
}

outlier_cols   <- c(predictors, config$response)
outlier_counts <- vapply(
  train_df[, outlier_cols],
  function(col) length(detect_outliers_iqr(col)),
  integer(1)
)

outlier_table <- data.frame(
  Variable  = outlier_cols,
  Outliers  = as.integer(outlier_counts),
  Percent   = round(100 * unname(outlier_counts) / nrow(train_df), 1),
  row.names = NULL
)

kable(
  outlier_table,
  caption = paste("Training-set values flagged by the 1.5×IQR rule,",
                   "computed on the training split only.")
)

freq_outliers  <- outlier_counts[["frequency"]]
thick_outliers <- outlier_counts[["suction_side_displacement_thickness"]]
angle_outliers <- outlier_counts[["attack_angle"]]
resp_outliers  <- outlier_counts[[config$response]]
```

Outliers were screened with the standard $1.5\times\text{IQR}$ fence, applied to
each variable using **training-set quantiles only** so that no information from
the held-out test rows enters the check. `chord_length` and
`free_stream_velocity` contribute zero flagged points, because - as noted in
1A - each takes only a handful of fixed values, so every observation sits well
inside its own IQR fence by construction. `frequency`, `attack_angle`, and
`suction_side_displacement_thickness` each contribute a nontrivial share
(`r freq_outliers`, `r angle_outliers`, and `r thick_outliers` training rows
respectively), and `r resp_outliers` response values are flagged.

These flagged points are **retained rather than removed**. The airfoil data
come from a controlled wind-tunnel experiment, not a survey or manual data-entry
process, so there is no mechanism here for typographical or transcription
errors; every extreme value is a physically valid test condition (a high
frequency band, a large angle of attack, a thin boundary layer) rather than a
data-quality defect. The IQR rule is doing exactly what it is designed to do on
a right-skewed engineering variable: flagging the upper tail of a real physical
distribution, not finding mistakes. Dropping these rows would bias the training
sample away from precisely the loud, high-frequency regime the model is meant to
predict, so all `r nrow(train_df)` training rows are kept for model fitting in
Problem 2.
```{r transform-check, echo=FALSE, warning=FALSE, message=FALSE}
skewness <- function(x) {
  z <- x - mean(x)
  mean(z^3) / (mean(z^2))^1.5
}

skew_raw <- vapply(train_df[, predictors], skewness, numeric(1))
skew_log <- c(
  frequency = skewness(log(train_df$frequency)),
  suction_side_displacement_thickness =
    skewness(log(train_df$suction_side_displacement_thickness))
)

skew_table <- data.frame(
  Variable           = names(skew_raw),
  Skewness           = unname(round(skew_raw, 3)),
  Skewness_after_log = c(
    round(skew_log[["frequency"]], 3), NA, NA,
    round(skew_log[["suction_side_displacement_thickness"]], 3), NA
  ),
  row.names = NULL
)

ols_raw_form <- reformulate(predictors, response = config$response)
ols_raw_fit  <- lm(ols_raw_form, data = train_df)

train_df_log <- train_df
train_df_log$frequency <- log(train_df_log$frequency)
train_df_log$suction_side_displacement_thickness <-
  log(train_df_log$suction_side_displacement_thickness)
ols_log_fit <- lm(ols_raw_form, data = train_df_log)

r2_raw <- summary(ols_raw_fit)$r.squared
r2_log <- summary(ols_log_fit)$r.squared

# Is the frequency effect itself curved rather than log-shaped?
r2_freq_linear <- summary(lm(train_df[[config$response]] ~ train_df$frequency))$r.squared
r2_freq_quad   <- summary(lm(train_df[[config$response]] ~ train_df$frequency +
                                I(train_df$frequency^2)))$r.squared
```

### Data cleaning summary
```{r cleaning-summary, echo=FALSE, warning=FALSE, message=FALSE}
cleaning_summary <- data.frame(
  Check = c("Missing values", "Duplicate rows", "Invalid values",
            "IQR outliers", "Skewness"),
  Result = c(
    paste(n_missing_total, "cells"),
    paste(n_duplicate_rows, "rows"),
    paste(sum(implausible_checks$Violations), "violations"),
    paste(sum(outlier_counts), "variable-level flags (a row may be flagged on more than one variable)"),
    "frequency and suction_side_displacement_thickness right-skewed"
  ),
  Decision = c(
    "No imputation needed",
    "No de-duplication needed",
    "No rows dropped",
    "Retained (valid experimental conditions, not errors)",
    "No log transform applied (does not improve the linear fit)"
  )
)

kable(
  cleaning_summary,
  caption = "Summary of data-cleaning checks and decisions carried into Problem 2."
)
```

## Data exploration
### Pairwise correlation among predictors
```{r eda-cor, echo=FALSE, warning=FALSE, message=FALSE}
# Pairwise correlation among predictors
cor_matrix <- cor(
  train_df[, predictors],
  use = "complete.obs"
)

# knitr::kable(
  #round(cor_matrix, 2),
  #caption = "Pairwise correlation matrix among predictors"
#)
```

```{r eda-heatmap, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.align="center"}
# Correlation heatmap
layout(matrix(c(1, 2), nrow = 1), widths = c(8.5, 1.5))
par(mar = c(5, 5, 2, 1))
cor_matrix <- cor(train_df[, predictors], use = "complete.obs")

image(1:ncol(cor_matrix), 1:ncol(cor_matrix), cor_matrix, 
      axes = FALSE, xlab = "", ylab = "", 
      col = heat.colors(1000, rev=TRUE), main = "Correlation Heatmap of Predictors")

axis(1, at = 1:ncol(cor_matrix), labels = colnames(cor_matrix), las = 2)
axis(2, at = 1:ncol(cor_matrix), labels = colnames(cor_matrix), las = 1)

par(mar = c(5, 1, 2, 3)) 
legend_scale <- matrix(seq(0, 1, length.out = 100), nrow = 1)
image(1, seq(0, 1, length.out = 100), legend_scale, 
      col = heat.colors(20, rev=TRUE), axes = FALSE, xlab = "", ylab = "")

axis(4, at = seq(0, 1, by = 0.1), las = 1, cex.axis = 0.8)
```

```{r eda-cor-summary, echo=FALSE, warning=FALSE, message=FALSE}
cor_off_diag <- cor_matrix
diag(cor_off_diag) <- NA
max_pair_idx <- which(abs(cor_off_diag) == max(abs(cor_off_diag), na.rm = TRUE), arr.ind = TRUE)[1, ]
max_pair_vars <- rownames(cor_matrix)[max_pair_idx]
max_pair_val  <- cor_matrix[max_pair_idx[1], max_pair_idx[2]]
```

Predictor-predictor correlations are weak across most pairs. The one exception
is `r max_pair_vars[1]` and `r max_pair_vars[2]`, correlated at
$r \approx `r round(max_pair_val, 2)`$ - a moderate relationship, plausible
given both describe the airfoil's boundary-layer geometry at the trailing edge.
No other pair exceeds $|r| \approx 0.5$. This single moderate correlation is
revisited with the VIF check below and is the main multicollinearity signal
this dataset offers ridge and lasso a reason to address in Problem 2.

### Response Distribution

```{r distribution-summary, echo=FALSE, warning=FALSE, message=FALSE}

hist(y_train)
summary <- as.matrix(summary(y_train))

kable(
  data.frame(
    Min. = summary["Min.", 1],
    `1st Qu.` = summary["1st Qu.", 1],
    Median = summary["Median", 1],
    Mean = summary["Mean", 1],
    `3rd Qu.` = summary["3rd Qu.", 1],
    Max. = summary["Max.", 1]
  ),
  digits = 2,
  align = "c"
)

resp_skew <- skewness(y_train)
```

The response ranges from `r round(min(y_train), 1)` to `r round(max(y_train), 1)`
dB with a median of `r round(median(y_train), 1)` dB and mean
`r round(mean(y_train), 1)` dB. The histogram and the small gap between mean and
median show a mild left skew (skewness $\approx `r round(resp_skew, 2)`$): most
measurements cluster in the `r round(quantile(y_train, 0.25), 0)`-
`r round(quantile(y_train, 0.75), 0)` dB range, with a longer tail toward the
quieter (lower-dB) end rather than the loud end. The skew is mild enough that no
response transformation is applied - RMSE and MAE on the original dB scale
remain directly interpretable as sound-pressure prediction errors, which matters
for Problem 1's evaluation step.

### Predictor Histogram  & Boxplot

1. Histogram

```{r eda-hist, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=1.4}

par(
  mfrow = c(1, 5),
  mar = c(2.5, 2, 1, 0.5)
)

for (var in predictors) {
  hist(
    train_df[[var]],
    main = "",
    xlab = var,
    breaks = 15
  )
}

par(mfrow = c(1,1))
```


2. Boxplot

```{r eda-boxplot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=3}
par(mfrow = c(1,5))
for (var in predictors) {
  boxplot(
    train_df[[var]],
    main = var
  )
}
par(mfrow = c(1,1))


```

The histograms and boxplots make the distinct-value counts visible. `chord_length`
and `free_stream_velocity` appear as `r n_levels_chord` and `r n_levels_vel`
narrow, evenly spaced spikes rather than a smooth spread, and `frequency` shows
`r n_levels_freq` spikes concentrated at the low end with a long right tail -
these are the fixed experimental levels described in 1A, not gaps in the data.
`attack_angle` and `suction_side_displacement_thickness` are closer to
continuous, right-skewed shapes, matching the skewness values quantified in 1B.
The boxplots' upper whiskers and flagged points for these two variables are the
same training-set values already counted in the 1B outlier check, retained for
the same reason: they are valid high-frequency / high-angle test conditions, not
data errors.

### Correlation between predictors and response

```{r eda-cor-res-pred, echo=FALSE, warning=FALSE, message=FALSE}
cor_response <- cor(
  train_df[, predictors],
  train_df[[config$response]]
)
colnames(cor_response) <- config$response

knitr::kable(
  round(cor_response, 2),
  caption = "Correlation between predictors and scaled_sound_pressure"
)

strongest_pred <- rownames(cor_response)[which.max(abs(cor_response[, 1]))]
strongest_cor  <- cor_response[strongest_pred, 1]
```

Every predictor's marginal correlation with the response is weak in magnitude
(all $|r| < 0.4$). `r strongest_pred` is the strongest single linear predictor
($r \approx `r round(strongest_cor, 2)`$), followed by
`suction_side_displacement_thickness`; `free_stream_velocity` is weakest and the
only predictor positively correlated with the response. No individual predictor
comes close to explaining the response on its own, which is consistent with the
assignment brief's description of this dataset as having "few features but
nonlinear relationships" - a linear combination of all five predictors, rather
than any single one, is needed, and Problem 2's regularized models are expected
to combine these weak individual signals rather than select one dominant
predictor and discard the rest.

# OLS, Ridge, and Lasso

## OLS baseline
```{r ols-fit, echo=FALSE}
# Convert x_train (matrix) and y_train into a data frame for lm()
train_data <- data.frame(y = y_train, x_train)

# Fit ordinary least squares on the standardized training set
ols_fit <- lm(y ~ ., data = train_data)


```

```{r ols-output, echo=FALSE, warning=FALSE, message=FALSE}
# Tidy coefficient table (term, estimate, std.error, statistic, p.value)
ols_coefs <- broom::tidy(ols_fit)

kable(
  ols_coefs,
  digits = 4,
  caption = "OLS coefficient estimates on standardized training data."
)
```

All estimates sit on the standardized scale fixed in 1B, so their magnitudes are
directly comparable across rows and the intercept is the training mean of
`scaled_sound_pressure`. A slope whose sign contradicts the predictor's marginal
correlation with the response is the first symptom of collinearity,
examined below.

```{r ols-metrics, echo=FALSE}
# In-sample fit quality
ols_summary <- summary(ols_fit)

# Training error (optimistic: the model has already seen these rows)
ols_train_pred    <- predict(ols_fit, newdata = train_data)
ols_train_metrics <- score_regression(y = y_train, prediction = ols_train_pred)

# Five-fold CV using the SAME foldid later passed to ridge and lasso.
# lm() has no built-in CV, so refit on each four-fold training set and
# score the held-out fold.

K <- max(foldid)
cv_mse_folds <- numeric(K)
for (k in seq_len(K)) {
  idx_train_k <- which(foldid != k)
  idx_val_k   <- which(foldid == k)
  fold_data   <- data.frame(y = y_train[idx_train_k], x_train[idx_train_k, , drop = FALSE])
  fold_fit    <- lm(y ~ ., data = fold_data)
  fold_pred   <- predict(fold_fit,
                         newdata = data.frame(x_train[idx_val_k, , drop = FALSE]))
  cv_mse_folds[k] <- mean((y_train[idx_val_k] - fold_pred)^2)
}
ols_cv_mse  <- mean(cv_mse_folds)
ols_cv_rmse <- sqrt(ols_cv_mse)

# Standard error of the CV-MSE across folds. cv.glmnet reports the same quantity
# as `cvsd` for ridge and lasso, so 2D can compare all three on equal footing.
ols_cv_se <- sd(cv_mse_folds) / sqrt(K)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Collect the fit and error numbers into one table
ols_fit_summary <- tibble::tibble(
  Quantity = c("R-squared", "Adjusted R-squared",
               "Training RMSE", "Training MAE",
               "5-fold CV RMSE", "5-fold CV MSE",
               "5-fold CV SE (of MSE)"),
  Value = c(ols_summary$r.squared, ols_summary$adj.r.squared,
            ols_train_metrics[["RMSE"]], ols_train_metrics[["MAE"]],
            ols_cv_rmse, ols_cv_mse, ols_cv_se)
)
kable(
  ols_fit_summary,
  digits = 4,
  caption = "OLS in-sample fit and five-fold cross-validation error on the shared folds."
)
```

The table gathers the in-sample fit and the five-fold cross-validation error on
the shared folds. The training rows are optimistic because the model was scored
on the same rows it was fitted on, whereas the CV rows estimate error on
unseen rows. The final row - the standard error of the CV-MSE across folds -
is the quantity `cv.glmnet` reports as `cvsd` for ridge and lasso, so 2D can
compare all three methods with error bars rather than bare point estimates.

```{r ols-helpers, echo=FALSE}
# Helper functions to extract coefficient and p-value by term name
ols_beta <- function(term) ols_coefs$estimate[ols_coefs$term == term]
ols_pval <- function(term) ols_coefs$p.value[ols_coefs$term == term]

# Correlation between the one moderately related predictor pair flagged in 1C
# (used in narrative below).
cor_angle_thick <- cor(x_train[, "attack_angle"], x_train[, "suction_side_displacement_thickness"])

# Quantify collinearity from this fit. VIF is scale-invariant, so standardizing
# in 1B leaves it unchanged and these values match the VIF table reported in 1C.
ols_vif      <- car::vif(ols_fit)
max_vif      <- max(ols_vif)
max_vif_term <- names(which.max(ols_vif))

# Marginal (1C) vs partial (this fit) effect of suction_side_displacement_thickness,
# used to illustrate how sharing variance with attack_angle changes its coefficient.
thick_marginal_cor <- cor(x_train[, "suction_side_displacement_thickness"], y_train)
```

**Model summary.** The OLS model is fitted on the `r nrow(x_train)` standardized
training observations using all `r ncol(x_train)` candidate predictors, with no
penalty and no variable selection. It is therefore the unregularized baseline
against which ridge and lasso are judged.

**Coefficient interpretation.** All five predictors are highly significant
($p < 0.001$ for every one, table above), so the ranking below is by
coefficient magnitude rather than by which predictors "matter".
- `frequency` carries the largest standardized coefficient
  ($\hat{\beta} \approx `r round(ols_beta("frequency"), 2)`$), confirming it as
  the strongest single linear driver of the response, consistent with it also
  having the strongest marginal correlation.
- `chord_length` ($\hat{\beta} \approx `r round(ols_beta("chord_length"), 2)`$)
  and `attack_angle`
  ($\hat{\beta} \approx `r round(ols_beta("attack_angle"), 2)`$) are the next
  largest contributors, both negative like `frequency`.
- `free_stream_velocity` ($\hat{\beta} \approx `r round(ols_beta("free_stream_velocity"), 2)`$)
  is the only predictor with a **positive** coefficient: higher free-stream
  velocity is associated with a higher (louder) scaled sound pressure once the
  other four predictors are held fixed.
- `suction_side_displacement_thickness`
  ($\hat{\beta} \approx `r round(ols_beta("suction_side_displacement_thickness"), 2)`$)
  has the smallest coefficient magnitude of the five, despite being the
  second-strongest predictor on its own($r \approx `r round(thick_marginal_cor, 2)`$)

```{r ols-qq, echo=FALSE,  fig.pos="H", out.extra="", fig.width=3.5, fig.height=3, fig.align="center"}

ols_std_resid <- rstandard(ols_fit)

qq <- qqnorm(ols_std_resid, plot.it = FALSE)

plot(qq$x, qq$y,
     type = "n",
     main = "Normal Q-Q - OLS standardized residuals",
     xlab = "Theoretical quantiles",
     ylab = "Standardized residuals")

grid(col = "grey88", lty = 1, lwd = 0.6)

abline(0, 1, col = "firebrick", lwd = 1.5, lty = 2)

points(qq$x, qq$y,
       pch = 19,
       col = "steelblue",
       cex = 0.8)

worst_resid <- order(abs(ols_std_resid), decreasing = TRUE)[1:3]

text(
  qq$x[worst_resid],
  qq$y[worst_resid],
  labels = train_id[worst_resid],
  pos = ifelse(qq$x[worst_resid] > 0, 2, 4),
  cex = 0.75,
  col = "firebrick"
)

shapiro_ols    <- shapiro.test(ols_std_resid)
big_resid_rows <- train_id[which(abs(ols_std_resid) > 2)]
max_qq_gap     <- max(abs(qq$y - qq$x))
```

**Residual normality.** The p-values quoted above are $t$-tests, and they lean
on the errors being approximately normal, so the assumption is checked rather
than taken on trust. Each point in the Q-Q plot is one wind-tunnel measurement:
its standardized residual against the value a normal sample of this size would
be expected to produce at that rank. Visually the points sit close to the
45-degree line - no residual departs from it by more than
`r round(max_qq_gap, 2)` standard deviations - but with
`r nrow(x_train)` training rows a Shapiro-Wilk test has enough power to detect
even small departures, and here it does
($W = `r round(shapiro_ols$statistic, 3)`$,
$p = `r round(shapiro_ols$p.value, 4)`$): formally, normality is rejected at the
5% level. $W$ itself is close to 1, so the rejection reflects a large sample
picking up a mild, not a severe, departure. `r length(big_resid_rows)`
residuals fall beyond two standard deviations, against roughly
`r round(0.05 * nrow(x_train))` expected by chance in a sample this size - a
noticeable excess, consistent with slightly heavier tails than a normal
distribution rather than a handful of extreme outliers.

Two consequences follow. The coefficient p-values in the table above should be
read as approximately, rather than exactly, valid, and because the tails are a
little heavier than normal, RMSE - which squares errors and so weights the
worst misses more heavily - is a somewhat more outlier-sensitive summary of
error here than it would be under strict normality; MAE in the same table is
the more robust companion figure for that reason.
## Ridge regression

### Cross-validation curve:
```{r ridge, echo = FALSE}
ridge <- analysis_model(x = x_train, y = y_train, alpha=0, foldid = foldid)
ridge_cvsd_min <- ridge$model$cvsd[ridge$model$lambda == ridge$model$lambda.min]
```

```{r ridge-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4, fig.height=3, fig.align="center"}
plot(ridge$model)
```
The mean cross-validation MSE decreases as $-Log(\lambda)$ increases (equivalently, as $\lambda$ decreases), suggesting that reducing the regularization strength improves predictive performance. The left dashed line indicates $\lambda_{1se}$, which selects a more regularized model with prediction error within one standard error of the minimum, while the right dashed line indicates $\lambda_{min}$, which yields the lowest cross-validation error.

### Lambda:
### Ridge penalty selection

```{r echo=FALSE, warning=FALSE, message=FALSE}

ols_cond <- safe_condition_numbers(x_train, lambda = 0)[2]

results <- data.frame(
  Metric = c(
    "Lambda",
    "MSE",
    "Condition number"
  ),
  `Lambda min` = c(
    ridge$lambda_min,
    ridge$mse_min,
    ridge$cond_min
  ),
  `Lambda 1se` = c(
    ridge$lambda_1se,
    ridge$mse_1se,
    ridge$cond_1se
  )
)

kable(
  results,
  digits = 4
)
```

With $\lambda_{min} = `r round(ridge$lambda_min, 4)`$, the 5-fold cross-validation MSE reaches its minimum value of 
$MSE_{min} = `r round(ridge$mse_min, 4)`$, indicating the best predictive performance among the candidate values of $\lambda$. The corresponding condition number, `r round(ridge$cond_min, 2)`, is already a substantial improvement over the unpenalized Gram matrix's `r round(ols_cond, 2)`. Even this mildly-collinear dataset (max VIF `r round(max_vif, 2)`) benefits measurably from a small ridge penalty.

The one-standard-error rule selects a larger penalty, $\lambda_{1se} = `r round(ridge$lambda_1se, 4)`$. This further reduces the condition number from `r round(ridge$cond_min, 2)` to `r round(ridge$cond_1se, 2)`, while the cross-validation MSE increases only
from `r round(ridge$mse_min, 3)` to `r round(ridge$mse_1se, 3)`, a difference of `r round(ridge$mse_1se - ridge$mse_min, 3)`. This difference is smaller than ridge's own cross-validation standard error at $\lambda_{min}$ (`r round(ridge_cvsd_min, 3)`, the cvsd reported by cv.glmnet, mirroring the per-fold SE computed for OLS). Thus, the two penalties are not statistically distinguishable on this evidence, making the additional stability from $\lambda_{1se}$ effectively free in terms of prediction
error.

### Coefficients:
```{r echo=FALSE, warning=FALSE, message=FALSE, echo = FALSE}

knitr::kable(
  data.frame(
    OLS = as.numeric(coef(ols_fit)),
    Lambda_Min = as.numeric(ridge$coef_min),
    Lambda_Max = as.numeric(ridge$coef_1se),
    row.names = names(ols_fit$coefficients)
  )
)
```

Compared with the OLS model, Ridge regression shrinks most regression coefficients toward zero, reducing their magnitudes through the regularization penalty, while retaining all five predictors in the model. `frequency`, `attack_angle`, `chord_length`, and `free_stream_velocity` all shrink exactly as expected, moving closer to zero as the penalty strengthens. `suction_side_displacement_thickness` is the one exception: its magnitude at $\lambda_{min}$ (`r round(ridge$coef_min["suction_side_displacement_thickness", ], 3)`) is slightly *larger* than its OLS estimate (`r round(ols_beta("suction_side_displacement_thickness"), 3)`), moving further from zero rather than shrinking. This is the same mechanism flagged in 2A: `suction_side_displacement_thickness` shares variance with the correlated `attack_angle` ($r \approx `r round(cor_angle_thick, 2)`$), and as Ridge shrinks `attack_angle`'s coefficient it redistributes some of that shared explanatory power back onto `suction_side_displacement_thickness`. Unlike a severe collinearity case, no sign changes anywhere in the table - the exception is a small redistribution, not an instability.

```{R echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=5, fig.align="center"}
n_vars <- nrow(ridge$model$glmnet.fit$beta)
plot(ridge$model$glmnet.fit)
legend("topleft", 
       legend = rownames(ridge$model$glmnet.fit$beta), 
       col = 1:n_vars, 
       lty = 1, 
       cex = 0.6)
```
As $\lambda$ increases, all five coefficients are progressively shrunk toward zero, demonstrating the regularization effect of Ridge regression; as $\lambda$ decreases, they move back toward their OLS estimates. `frequency` remains the largest-magnitude coefficient across the range spanned by $\lambda_{min}$ and $\lambda_{1se}$ (`r round(ridge$coef_min["frequency", ], 2)` to `r round(ridge$coef_1se["frequency", ], 2)`), consistent with it being the strongest predictor throughout 1C-2A. `free_stream_velocity` is the only coefficient that stays positive along the entire path, mirroring the sign pattern already seen in OLS. No coefficient crosses zero or reverses sign between $\lambda_{min}$ and $\lambda_{1se}$ - the ridge penalty here is redistributing magnitude among correlated predictors (as seen above for `suction_side_displacement_thickness`), not fighting an unstable or sign-flipping fit.

## Lasso regression

### Model fitting and cross-validation
```{r lasso-fit, echo = FALSE}
lasso_cv <- analysis_model(
    x = x_train,
    y = y_train,
    alpha = 1,
    foldid = foldid
)
```

```{r lasso-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4, fig.height=3, fig.align="center"}
plot(lasso_cv$model)

```

Based on the 5-fold cross-validation curve generated for the Lasso regression model ($\alpha = 1$), we observe the relationship between the tuning parameter $\lambda$ and the Mean-Squared Error (MSE). The top axis indicates the number of active (non-zero) predictors retained in the model at each level of penalization.

* The Cross-Validation Curve: The red dots represent the average CV-MSE across the 5 folds for each $\lambda$ value, with the gray error bars denoting one standard error above and below the mean. As $-\log(\lambda)$ increases (moving from left to right, meaning the penalty weakens), the MSE initially drops, reaches a minimum, and then slightly stabilizes as the model approaches the unpenalized OLS limit.

* Optimal $\lambda$ ($\lambda_{\text{min}}$): The right vertical dashed line marks `lambda.min`
($-\log(\lambda) \approx `r round(-log(lasso_cv$lambda_min), 2)`$), the value that
strictly minimizes the cross-validation MSE. At this optimal point for
prediction accuracy the Lasso penalty retains all
`r lasso_cv$nonzero_min` predictors.

* Parsimonious $\lambda$ ($\lambda_{\text{1se}}$): The left vertical dashed line
marks `lambda.1se`
($-\log(\lambda) \approx `r round(-log(lasso_cv$lambda_1se), 2)`$), selected by
the one-standard-error rule. This value applies a stronger penalty to produce a
more conservative model while keeping the error within one standard deviation
of the absolute minimum MSE; whether it also produces a *sparser* model, in the
sense of dropping predictors, is checked directly below rather than assumed.

### Analysis of $\lambda_{\text{min}}$

```{r lasso-min, echo=FALSE, warning=FALSE, message=FALSE}

results <- data.frame(
  Metric = c("Lambda min", "MSE", "Condition number"),
  Value = c(
    lasso_cv$lambda_min,
    lasso_cv$mse_min,
    lasso_cv$cond_min
  )
)

kable(
  results,
  digits = 4
)
```
The optimal tuning parameter selected by five-fold cross-validation is
$\lambda_{\text{min}} \approx `r round(lasso_cv$lambda_min, 4)`$. This value
strictly minimizes the cross-validation Mean Squared Error to
$`r round(lasso_cv$mse_min, 3)`$, essentially matching OLS's own CV-MSE
(`r round(ols_cv_mse, 3)`, 2A) - unsurprising given how small $\lambda_{min}$
is. The condition number at this threshold, `r round(lasso_cv$cond_min, 2)`,
is a modest improvement over the unpenalized Gram matrix
(`r round(ols_cond, 2)`, 1C/3C).

At this level of regularization, the Lasso model retains all
`r lasso_cv$nonzero_min` predictors - `frequency`, `attack_angle`,
`chord_length`, `free_stream_velocity`, and
`suction_side_displacement_thickness` - none is shrunk exactly to zero. This is
consistent with the OLS and VIF results in 2A: every predictor was already
highly significant with only mild shared variance, so there is no clearly
redundant predictor for Lasso's $L_1$ penalty to remove at this tuning value.

The retained coefficients preserve the same ranking seen in OLS and ridge:
`frequency` has the largest magnitude
(`r round(lasso_cv$coef_min["frequency", ], 3)`), followed by `chord_length`
(`r round(lasso_cv$coef_min["chord_length", ], 3)`) and `attack_angle`
(`r round(lasso_cv$coef_min["attack_angle", ], 3)`), with
`free_stream_velocity` (`r round(lasso_cv$coef_min["free_stream_velocity", ], 3)`)
the only positive coefficient and
`suction_side_displacement_thickness`
(`r round(lasso_cv$coef_min["suction_side_displacement_thickness", ], 3)`)
the smallest in magnitude.

Overall, $\lambda_{\text{min}}$ produces a model whose cross-validation error is
essentially indistinguishable from OLS's while retaining every predictor: on
this dataset, Lasso's contribution at $\lambda_{min}$ is confirming that all
five predictors carry real signal, rather than selecting a subset of them.


### Analysis of $\lambda_{\text{1se}}$

```{r lasso-1se, echo=FALSE, warning=FALSE, message=FALSE}

results <- data.frame(
  Metric = c("Lambda 1se", "MSE", "Condition number"),
  Value = c(
    lasso_cv$lambda_1se,
    lasso_cv$mse_1se,
    lasso_cv$cond_1se
  )
)

kable(
  results,
  digits = 4
)
```
The one-standard-error rule selects
$\lambda_{\text{1se}} \approx `r round(lasso_cv$lambda_1se, 4)`$, which applies
a stronger regularization penalty than $\lambda_{\text{min}}$. The
corresponding five-fold cross-validation MSE is
$`r round(lasso_cv$mse_1se, 3)`$. Although this is higher than the minimum
achieved by $\lambda_{\text{min}}$, it remains within one standard error of
that minimum by construction. The stronger penalty reduces the condition number
to `r round(lasso_cv$cond_1se, 2)`, a substantial further improvement in
conditioning.

Under $\lambda_{\text{1se}}$, the Lasso model still retains all
`r lasso_cv$nonzero_1se` predictors - unlike the sparser model this rule often
produces, no predictor is driven to zero even under this stronger penalty.
Coefficient magnitudes shrink further relative to $\lambda_{min}$ (for example
`frequency` falls from `r round(lasso_cv$coef_min["frequency", ], 3)` to
`r round(lasso_cv$coef_1se["frequency", ], 3)`), but the set of active
predictors is unchanged. Combined with the $\lambda_{min}$ result above, this
indicates that on this dataset Lasso's embedded feature-selection mechanism
finds nothing to prune: every predictor's contribution is large enough,
relative to the cross-validated penalty, to survive even the more conservative
tuning rule.


### Coefficient Path

```{r lasso-coef-path, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=5, fig.align="center"}

plot(lasso_cv$model$glmnet.fit, xvar = "lambda", label = TRUE)

legend("topleft", 
       legend = rownames(lasso_cv$model$glmnet.fit$beta), 
       col = 1:nrow(lasso_cv$model$glmnet.fit$beta), 
       lty = 1, 
       cex = 0.7,
       lwd = 2,
       title = "Predictors")

```


Overall, the coefficient path shows that as $-\log(\lambda)$ increases (equivalently, as $\lambda$ decreases), the regularization penalty becomes weaker and coefficients grow in magnitude toward their OLS values. Under strong enough regularization all five paths do eventually collapse to zero at the far left of the plot, but by the range spanned by $\lambda_{min}$ and $\lambda_{1se}$ - the only penalty strengths this analysis tunes to - every predictor is already active, matching the retained-predictor counts reported above.

`frequency` enters first (i.e. survives down to the smallest $\lambda$) and consistently has the largest-magnitude coefficient along the path, confirming it as the most influential predictor. `chord_length` and `attack_angle` follow a similar trajectory at smaller magnitudes, and `free_stream_velocity` is the only path that stays on the positive side of zero throughout. `suction_side_displacement_thickness` tracks a comparatively flat, small-magnitude path across the region shown - the same predictor whose coefficient 2B found to move *against* the general shrinkage trend under ridge.

## Controlled comparison and locked core model

```{r core-comparison, echo=FALSE, warning=FALSE, message=FALSE}
# Build a comparison from training/CV quantities only.
ols_summary <- data.frame(
  tunning_rule = "None",
  lambda = 0,
  cv_error = ols_cv_mse,
  cond = safe_condition_numbers(x_train, lambda = 0)[2],
  coefficient_norm = sqrt(sum(coef(ols_fit)[-1]^2)),
  num_non_zero = sum(coef(ols_fit) != 0, na.rm = TRUE) - 1,
  row.names = "OLS"
)

ridge_summary <- data.frame(
  tunning_rule = c("lambda.min", "lambda.1se"),
  lambda = c(ridge$lambda_min, ridge$lambda_1se),
  cv_error = c(ridge$mse_min, ridge$mse_1se),
  cond = c(ridge$cond_min, ridge$cond_1se),
  coefficient_norm = c(sqrt(sum(ridge$coef_min[-1]^2)),sqrt(sum(ridge$coef_1se[-1]^2))),
  num_non_zero = c(ridge$nonzero_min,ridge$nonzero_1se),
  row.names = c("Ridge"," ")
)
lasso_summary <- data.frame(
  tunning_rule = c("lambda.min", "lambda.1se"),
  lambda = c(lasso_cv$lambda_min, lasso_cv$lambda_1se),
  cv_error = c(lasso_cv$mse_min, lasso_cv$mse_1se),
  cond = c(lasso_cv$cond_min, lasso_cv$cond_1se),
  coefficient_norm = c(sqrt(sum(lasso_cv$coef_min[-1]^2)),sqrt(sum(lasso_cv$coef_1se[-1]^2))),
  num_non_zero = c(lasso_cv$nonzero_min, lasso_cv$nonzero_1se),
  row.names = c("Lasso", "  ")
)

knitr::kable(rbind(ols_summary,ridge_summary,lasso_summary))

```

The training comparison differs from the classic “lasso selects, ridge
stabilizes” narrative. All five candidates-OLS, ridge, and lasso under both
tuning rules-retain all `r ncol(x_train)` predictors: lasso performs no
variable selection at either $\lambda_{min}$ or $\lambda_{1se}$, so it provides
no interpretability advantage over ridge. Their CV errors are also very
similar, ranging from
`r round(min(rbind(ols_summary,ridge_summary,lasso_summary)$cv_error), 2)` to
`r round(max(rbind(ols_summary,ridge_summary,lasso_summary)$cv_error), 2)`,
within one CV standard error of each other (`r round(ols_cv_se, 2)`).
Thus, none is distinguishable by predictive accuracy alone.

Conditioning provides the main distinction: ridge at $\lambda_{min}$ nearly
matches OLS and lasso in CV error while roughly halving the condition number
(`r round(ridge$cond_min, 2)` vs. OLS's `r round(ols_cond, 2)`), while
$\lambda_{1se}$ further improves conditioning for both ridge and lasso at a
small additional CV-error cost. Since lasso does not sparsify this model, the
usual accuracy–interpretability trade-off does not apply; the relevant
trade-off is prediction accuracy versus numerical stability. On this basis,
**ridge at $\lambda_{min}$** is nominated as the core candidate: it is
statistically tied with OLS and lasso in training CV error while providing a
better-conditioned fit. This recommendation uses only the training data and
will be evaluated later on the locked holdout set.

The $\lambda_{min}$ and $\lambda_{1se}$ rules reflect different objectives.
$\lambda_{min}$ minimizes CV error and prioritizes predictive accuracy,
whereas $\lambda_{1se}$ selects the largest $\lambda$ within one standard
error of the minimum, favouring stronger regularization and greater stability
at comparable predictive performance. Thus, CV error assesses predictive
performance, variable retention assesses selection, and interpretability
concerns how easily the model can be understood. In our results, the trade-off
appears in *conditioning* rather than *sparsity*: $\lambda_{1se}$ retains the
same number of predictors as $\lambda_{min}$ for both ridge and lasso,
but accepts a small increase in CV error for a meaningfully better-conditioned
fit. This is therefore a stability–accuracy trade-off rather than a
complexity–accuracy trade-off.

# Experimental Design

## Dataset selection

### Data Overview

```{r data-configuration-anova, echo = FALSE}

data <- read.csv("data/penguins_size.csv", header = TRUE)

# Treat "." as a missing value
data[data == "."] <- NA

config <- list(
  file = "penguins_size.csv",
  response = "body_mass_g",
  factors = c("species", "sex")
)

data_raw <- data

analysis_data <- data_raw[
  ,
  c(config$response, config$factors),
  drop = FALSE
]

analysis_data <- na.omit(analysis_data)

if (anyNA(analysis_data)) {
  stop("Missing values remain in the analysis data.")
}
```

```{r data-overview-anova, echo=FALSE, warning=FALSE, message=FALSE}
#str(analysis_data)

#table(analysis_data$species)
#table(analysis_data$sex)

#table(
#  analysis_data$species,
#  analysis_data$sex
#)

data_overview <- tibble(
  Variable = names(data_raw),
  Role = ifelse(
    names(data_raw) == config$response,
    "Response",
    "Predictor"
  ),
  Type = vapply(
    data_raw,
    function(x) class(x)[1L],
    character(1)
  ),
  Distinct = vapply(
    data_raw,
    dplyr::n_distinct,
    integer(1),
    na.rm = TRUE
  ),
  Missing = vapply(
    data_raw,
    function(x) sum(is.na(x)),
    integer(1)
  )
)

kable(
  data_overview,
  caption = "Data Overview Before Missing-Value Removal"
)
# n_duplicate_rows <- sum(duplicated(analysis_data))

# n_duplicate_rows
```

For 2-Way ANOVA, this method requires the dataset to have at least 2 categorical variables along with a continuous response variable for the analysis.

In general, from our chosen dataset, we decide to evaluate the factors that are associated with penguin body mass. This dataset contains information about 344 penguins, including species, island, physical measurements such as culmen length, culmen depth, and flipper length, body mass, and sex. It is suitable for exploratory data analysis and 2-Way ANOVA analysis based on the biological and physical characteristics of penguins.

Specifically, the dataset contains 3 categorical variables: `species` (Adelie/Chinstrap/Gentoo), `island` (Torgersen/Biscoe/Dream), and `sex` (MALE/FEMALE). In this method, we select `species` and `sex` as the two categorical factors and `body_mass_g` as the continuous response variable. The analysis is used to test whether the two factors are associated with differences in mean body mass, as well as whether there is an interaction between the two factors.
  
### Research Questions & Hypotheses

We investigate whether penguin species and sex are associated with differences in mean body mass, and whether there is an interaction between the two factors.

We follow the 2-way ANOVA model:


$$
Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}
$$

with $\epsilon_{ijk} \sim N(0,\sigma^2)$.

- $\mu$ is the overall mean.
- $\alpha_i$ is the effect of the $i$-th level of Factor A, `species`.
- $\beta_j$ is the effect of the $j$-th level of Factor B, `sex`.
- $(\alpha\beta)_{ij}$ is the interaction effect between Factor A and Factor B.
- $\epsilon_{ijk}$ is the random error term.

#### Factor A: Penguin Species

**Research Question:**  
Is penguin species associated with differences in mean body mass?

\[
\begin{cases}
H_{0a}: \alpha_1 = \alpha_2 =  \alpha_3 = 0\\
H_{1a}: \text{Not all } \alpha_i \text{ are equal to }0.
\end{cases}
\]

#### Factor B: Penguin Sex

**Research Question:**  
Is the sex of penguin associated with differences in mean body mass?

\[
\begin{cases}
H_{0b}: \beta_1 = \beta_2 = 0\\
H_{1b}: \text{Not all } \beta_j \text{ are equal to } 0.
\end{cases}
\]

#### Interaction Effect

**Research Question:**  
Does the association between sex and penguin body mass depend on the penguin species?

\[
\begin{cases}
H_{0ab}: (\alpha\beta)_{ij} = 0 \quad \text{for all } i,j\\
H_{1ab}: \text{Not all } (\alpha\beta)_{ij} \text{ are equal to } 0.
\end{cases}
\]

### Descriptive Statistics

#### Data Summary Statistics
```{r data-response-summary-anova, echo=FALSE}

# Overall summary statistics on response
overall_summary <- analysis_data %>%
  summarise(
    N = n(),
    Mean = mean(body_mass_g, na.rm = TRUE),
    Median = median(body_mass_g, na.rm = TRUE),
    Variance = var(body_mass_g, na.rm = TRUE),
    SD = sd(body_mass_g, na.rm = TRUE),
    Min = min(body_mass_g, na.rm = TRUE),
    Max = max(body_mass_g, na.rm = TRUE)
  )

overall_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    caption = "Overall Summary Statistics of Penguin Body Mass",
    align = "c"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    position = "center"
  )

```

The dataset contains 333 observations of penguin body mass after removing observations with missing values. The mean body mass is 4207.06 g, with a median of 4050 g, indicating that the body masses are centered around approximately 4.0 to 4.2 kg. The standard deviation is 805.22 g, showing moderate variability in body mass, while the observed values range from 2700 g to 6300 g. The mean being slightly higher than the median suggests a mild right skew in the overall response distribution.

```{r data-treatment-combination-summary, echo=FALSE}

# Summary statistics by treatment combination
group_summary <- analysis_data %>%
  group_by(
    species,
    sex
  ) %>%
  summarise(
    N = n(),
    Mean = mean(body_mass_g, na.rm = TRUE),
    Median = median(body_mass_g, na.rm = TRUE),
    Variance = var(body_mass_g, na.rm = TRUE),
    SD = sd(body_mass_g, na.rm = TRUE),
    Min = min(body_mass_g, na.rm = TRUE),
    Max = max(body_mass_g, na.rm = TRUE),
    .groups = "drop"
  )

group_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    caption = "Summary Statistics by Treatment Combination",
    col.names = c(
      "Species",
      "Sex",
      "N",
      "Mean",
      "Median",
      "Variance",
      "SD",
      "Min",
      "Max"
    ),
    align = "c"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    position = "center"
  )
```
The dataset contains three levels of species and two levels of sex. Male penguins generally have higher mean body mass than female penguins across all three species.

The difference in mean body mass between males and females varies across species. The difference is approximately 674.65 g for Adelie penguins, 411.76 g for Chinstrap penguins, and 805.10 g for Gentoo penguins.
This variation provides descriptive evidence of a potential interaction between species and sex.

Among the six treatment combinations, the Gentoo male group has the highest mean body mass (5484.84 g), while the Adelie female group has the lowest mean body mass (3368.84 g). The group variances range from
72,565.64 to 120,278.25, with the Adelie male group having the largest variance. Therefore, the homogeneity of variance assumption should be checked before conducting the two-way ANOVA.

#### Visualization

##### Histogram of Penguin Body Mass

```{r data-hist, echo=FALSE, fig.width=8, fig.height=3.5, out.width="90%", fig.align="center"}

library(ggplot2)
library(patchwork)

p1 <- ggplot(analysis_data, aes(x = body_mass_g)) +
  geom_histogram(
    bins = 20,
    color = "black"
  ) +
  labs(
    title = "Distribution of Penguin Body Mass",
    x = "Body Mass",
    y = "Frequency"
  ) +
  theme_minimal()

p2 <- ggplot(
  analysis_data,
  aes(
    x = interaction(species, sex, sep = " x "),
    y = body_mass_g
  )
) +
  geom_boxplot() +
  scale_x_discrete(
    labels = function(x) gsub(" x ", "\n", x)
  ) +
  labs(
    title = "Penguin Body Mass by Treatment Combination",
    x = "Treatment Combination",
    y = "Body Mass (g)"
  ) +
  theme_minimal()

p1 + p2
```

The histogram shows that penguin body masses range from approximately 2700 g to 6300 g, with most observations concentrated between 3000 g and 5000 g. The distribution shows a mild right skew, with a longer tail
toward higher body masses. This is consistent with the mean body mass (4207.06 g) being higher than the median (4050 g). The distribution also shows relatively high frequencies around 3500 to 4000 g, while fewer observations are found at the upper end of the body mass range.

##### Box Plot 

```{r data-boxplot, echo=FALSE, fig.width=6.5, fig.height=4.5, out.width="95%", fig.align="center"}

```

The box plot compares the distribution of penguin body mass across the six treatment combinations of species and sex. Male penguins generally have higher median and mean body mass than female penguins within each species.

The difference between male and female body mass appears across all three species. The largest difference is observed for Adelie penguins, with mean body mass increasing from 3368.84 g for females to 4043.49 g for males. For Chinstrap penguins, the corresponding means are 3527.21 g and 3938.97 g, while for Gentoo penguins they are 4679.74 g and 5484.84 g.

The spread of the groups is not completely uniform, with some groups showing wider variability than others. A few potential low outliers are also visible, particularly in the Chinstrap female group. Overall, the
plot suggests that both species and sex may be associated with penguin body mass, while the differences in the male-female effect across species provide descriptive evidence of a possible interaction between the two factors.

##### Interaction Plot

```{r interaction-plot, echo=FALSE, fig.width=6, fig.height=4, out.width="90%", fig.align="center"}

ggplot(
  analysis_data,
  aes(
    x = sex,
    y = body_mass_g,
    group = species,
    linetype = species
  )
) +
  stat_summary(
    fun = mean,
    geom = "line"
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 3
  ) +
  labs(
    title = "Interaction Between Species and Sex",
    x = "Sex",
    y = "Mean Body Mass (g)",
    linetype = "Species"
  ) +
  theme_minimal()
```

The interaction plot shows the mean body mass for each combination of species and sex. Across all three species, male penguins have higher mean body mass than female penguins. The mean body mass increases by
approximately 674.66 g for Adelie, 411.76 g for Chinstrap, and 805.10 g for Gentoo penguins.

The lines are not perfectly parallel. In particular, the Adelie and Chinstrap lines cross between the female and male groups, while the Adelie and Gentoo lines show relatively similar slopes. This provides
some descriptive evidence that the difference in body mass between males and females may vary across species.

However, the interaction pattern appears relatively modest overall, and the plot alone cannot determine whether the interaction is statistically significant. The significance of the species-by-sex interaction will be formally assessed using the two-way ANOVA.

#### Assumptions Check

##### Normality of Residuals

```{r fit-model, echo=FALSE}
# fit model for checking assumptions
model <- aov(
  body_mass_g ~
    species * sex,
  data = data
)

residual <- residuals(model)
```

```{r shapiro, echo = FALSE}
shapiro <- shapiro.test(residual)
kable(
  data.frame(
    W = shapiro$statistic,
    "p-value" = shapiro$p.value,
    row.names = ""
  ),
)
```

The normality assumption was assessed using the Shapiro-Wilk test. The hypotheses for
the Shapiro-Wilk test are:

\[
\begin{cases}
H_0: \text{The residuals follow a normal distribution}\\
H_1: \text{The residuals do not follow a normal distribution}
\end{cases}
\]


The Shapiro-Wilk test resulted in

$$
W = 0.99776,\qquad p = 0.9367.
$$

Since the p-value is greater than 0.05, we fail to reject $H_0$. Therefore, there is no statistical evidence that the ANOVA residuals deviate from a normal distribution. The Shapiro-Wilk test gives $W=0.99776$ with a p-value of $0.9367$, providing further support for the normality assumption.

Overall, the residuals show no substantial departure from normality, and the normality assumption is considered satisfied for the subsequent two-way ANOVA.

##### Homogeneity of Variances 

```{r homogeneity, echo = FALSE}
leveneTest(
  body_mass_g ~
    species * sex,
  data = analysis_data
)
```

The homogeneity of variance assumption was assessed using the treatment group box plots and Levene's test. The six treatment combinations are formed by `species` and `sex`.

The hypotheses are:


\[
\begin{cases}
H_0:
\sigma_1^2 = \sigma_2^2 = \sigma_3^2 = \cdots =\sigma_6^2\\
H_1:
\text{At least one treatment group has a different variance}
\end{cases}
\]

Levene's test produced

$$
F(5,327) = 1.3908,\qquad p = 0.2272
$$
Since the p-value is greater than 0.05, we fail to reject $H_0$. Therefore, there is no statistical evidence that the variances differ across the six treatment combinations.

The group variances range from approximately 72,565.64 to 120,278.25, with the Adelie male group having the largest variance and the Adelie female group having the smallest variance. Although the group variances
are not identical, Levene's test suggests that the differences are not statistically significant. Therefore, the homogeneity of variance assumption is considered satisfied for the subsequent two-way ANOVA.

##### Independence of Errors

```{r independence-check, echo = FALSE, eval = FALSE}
sum(duplicated(analysis_data))
```

The independence assumption requires that the error terms are independent:

$$
\epsilon_{ijk} \overset{ind}{\sim} N(0,\sigma^2).
$$

The independence assumption was assessed based on the structure of the dataset. Each observation represents an individual penguin, and the dataset does not contain repeated measurements or a time-dependent
structure. Although some observations have identical values for the selected variables (`body_mass_g`, `species`, and `sex`), these do not necessarily represent duplicate penguins, since different penguins can
naturally have the same body mass and characteristics.

Therefore, independence is considered reasonable based on the available dataset information. However, it cannot be formally verified because the dataset does not provide a unique identifier or detailed information about the sampling procedure.

#### Overall Assessment

The independence assumption is considered reasonable based on the structure of the dataset. Each observation represents an individual penguin, and there is no time-dependent or repeated-measurement structure identified in the analysis data. However, independence cannot be formally verified because the dataset does not provide a unique identifier or detailed information about the sampling procedure.

The normality assumption is considered satisfied based on both the Q-Q plot and the Shapiro-Wilk test ($p=0.9367$), which provides no statistical evidence of a departure from normality. The homogeneity of variance assumption is also considered satisfied based on the non-significant Levene's test ($p=0.2272$).

Overall, the assumptions are considered reasonably adequate for conducting the two-way ANOVA. No substantial concerns are observed regarding normality or homogeneity of variance, while the independence assumption is considered reasonable based on the available dataset information.

## ANOVA / Factorial design analysis
```{r  echo = FALSE}
model <- aov(body_mass_g ~ species * sex ,data=analysis_data)
```

```{r echo=FALSE}
anova_table <- as.data.frame(summary(model)[[1]])
p_value <- anova_table$`Pr(>F)`
f_value <- anova_table$`F value`

kable(anova_table, digits=3, caption = "ANOVA result")
```
Using significant level $\alpha$ of 0.05, we observe that all these p-value are less than $\alpha$.

For species, the sum of squares is 145,190,219, which is the largest among all factors. Together with the very large F-value of 758.358 and p < 0.001, this indicates that species explains a very large amount of the variation in the response variable and has a highly significant effect.

For sex, the sum of squares is 37,090,262, which is also large compared with the residual sum of squares. The F-value of 387.460 and p < 0.001 show that sex has a highly significant effect on the response variable.

For the species - sex interaction, the degrees of freedom is 2, because the interaction involves the three species and two sex groups. Its sum of squares is 1,676,557, which is much smaller than the sums of squares for species and sex. However, the interaction is still statistically significant, with an F-value of 8.757 and p < 0.001. This indicates that although the interaction explains less variation, the effect of sex still differs significantly among species.
```{r echo=FALSE, fig.width = 4, fig.height = 3, fig.align='center'}
plot(model, which = 1)
```

This residuals are generally scattered around the horizontal zero line across the different fitted values, and the red trend line remains close to zero. This indicates that there is no strong systematic pattern or curvature in the residuals, suggesting that the ANOVA model captures the main structure of the data reasonably well.

The spread of the residuals is somewhat different between the fitted-value groups, but there is no clear funnel-shaped pattern where the spread consistently increases or decreases as the fitted values increase. Therefore, there is no strong visual evidence of severe heteroscedasticity from this plot.

```{r echo=FALSE, fig.width = 4, fig.height = 3, fig.align = "center"}
plot(model, which = 2)
```

This Q–Q plot shows most of the points closely follow the diagonal reference line, particularly in the middle section of the plot, indicating that the residuals are approximately normally distributed. This suggests that the assumption of normality required for ANOVA is reasonably satisfied.

## Posthoc analysis

```{r}
tukey <- TukeyHSD(model, which = "species")
```
```{r echo=FALSE}
kable(tukey$species, digits = 3, caption = "Tukey HSD Results")
```

The comparison between Chinstrap and Adelie is not statistically significant, with an adjusted p-value of 0.824, which is greater than 0.05. The confidence interval also includes zero (-80.026 to 133.874). Therefore, there is insufficient evidence to conclude that the mean response differs between Chinstrap and Adelie.

In contrast, Gentoo differs significantly from Adelie, with a mean difference of 1386.273 and an adjusted p-value less than 0.001. The 95% confidence interval (1296.307 to 1476.238) does not contain zero, providing strong evidence that the two species have different mean responses. Since the mean difference is positive, the mean response for Gentoo is higher than Adelie by approximately 1386.273 units.

Similarly, Gentoo differs significantly from Chinstrap, with a mean difference of 1359.349 and an adjusted p-value less than 0.001. The 95% confidence interval (1248.611 to 1470.087) is entirely positive, indicating that the mean response for Gentoo is higher than Chinstrap by approximately 1359.349 units.

Overall, the Tukey HSD results show that Gentoo is significantly different from both Adelie and Chinstrap, whereas Adelie and Chinstrap do not differ significantly from each other. This provides more specific information about the significant species effect identified by the ANOVA.

```{r}
sex_analysis <- t.test(body_mass_g ~ sex,data=analysis_data)

tidy_t_test <- tidy(sex_analysis)
```

```{r echo=FALSE}
kable(
  tidy_t_test[, c("estimate", "statistic", "p.value", "conf.low", "conf.high")],
  digits = 3,
  col.names = c("Mean Difference", "t-Statistic", "p-value", "Lower CI", "Upper CI"),
  caption = "Independent t-test Comparing Penguin Body Mass by Sex"
)
```

An independent-samples t-test was conducted to compare penguin body mass between male and female penguins. The results showed a mean difference of -683.412 grams, with a t-statistic of -8.555 and a p-value of < 0.001.

The very small p-value indicates a statistically significant difference in body mass between the two sexes. The mean difference of -683.412 grams indicates that the first sex group in the comparison has an average body mass approximately 683.4 grams lower than the second sex group.

The 95% confidence interval for the mean difference ranges from -840.578 to -526.245 grams. Since the entire interval is below zero, the estimated difference is consistently in the same direction. Therefore, the results indicate that one sex has substantially greater average body mass than the other.

Overall, sex is associated with a significant difference in penguin body mass, with an average difference of approximately 683 grams between the two sexes.
```{r, echo = FALSE}
emm_result <- emmeans(model, pairwise ~ sex | species, adjust = "tukey")

```
```{r echo=FALSE}
kable(
  as.data.frame(emm_result$contrasts), 
  digits = 3,
  col.names = c("Contrast", "Species", "Difference (g)", "Std. Error", "df", "t-ratio", "p.value"),
  align = "llrrrrr",          
  caption = "Pairwise Comparisons of Body Mass Between Sexes Stratified by Penguin Species"
)
```

Pairwise comparisons showed that male penguins had significantly higher body mass than females in all three species (p<0.001). The differences were approximately 674.7 g for Adelie, 411.8 g for Chinstrap, and 805.1 g for Gentoo. The largest sex difference occurred in Gentoo, while the smallest occurred in Chinstrap.

### Practical significance

The results are not only statistically significant but also practically meaningful. The differences in body mass between species are large, particularly between Gentoo and the other two species, with differences of more than 1,300 g. The sex differences are also substantial, ranging from approximately 412 g to 805 g, with males consistently heavier than females.

Therefore, species and sex should be considered when assessing or comparing penguin body mass. Ignoring these factors could lead to misleading comparisons, especially when comparing groups with different species or sex compositions.

## Conclusion and evaluation

The two-way ANOVA showed that species and sex both have significant effects on penguin body mass (p<0.001). The Species x Sex interaction was also significant (p<0.001), indicating that the effect of sex on body mass varies among species. Tukey's HSD showed that Gentoo differs significantly from both Adelie and Chinstrap, while Adelie and Chinstrap do not differ significantly. Pairwise comparisons further showed that males have significantly higher body mass than females in all three species, with the largest difference observed in Gentoo.

The analysis provides strong evidence of differences in body mass by species and sex. However, the ANOVA assumptions should be considered, particularly normality and homogeneity of variance. The data may also be unbalanced across groups, which can affect the analysis. Future analysis could use robust methods or transformations if assumption violations are substantial.

Based on these results, species and sex should both be considered when analyzing or predicting penguin body mass. In particular, the significant interaction suggests that the effect of sex should not be interpreted independently of species. Further studies could collect more balanced samples across species and sexes to improve the reliability and generalizability of the findings.