---
title: "ASESII 24A02 Final Project"
subtitle: "Predict noise level generated by NACA airfoil blade sections"
author: |
  | Group 02
  | Nguyen Thi Yen Nhi (24125071)
  | Nguyen Khang Thinh (24125104)
  | Nguyen Van Tinh (24125106)
  | Nguyen Le Quoc Huy (24125100)
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
link-citations: true
nocite: |
  @loshchilov2019decoupledweightdecayregularization, @krogh1992simple
params:
  group_number: 02
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere}
  - \usepackage{float}
  - \usepackage{longtable}
---




# Authorship and contribution statement {-}

<!-- We confirm that every group member can explain the submitted analysis. AI use is reported in `Group04_AI_Log.csv`, and the group checked every AI-assisted item used in this report. -->

```{=latex}
\begingroup
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{p{0.25\linewidth} p{0.09\linewidth} p{0.33\linewidth} p{0.24\linewidth}}
\hline
Student & ID & Main contributions & Verification performed \\
\hline
Nguyen Thi Yen Nhi & 24125071 &
1A (Data analysis), 2A (OLS), 4E (Question Explanation) &
Verify AI texts \\
\hline
Nguyen Khang Thinh & 24125104 &
1C (EDA), 2C (Lasso), 4B (Elastic Net), 4C (ReLU features), 4D (Holdout Eval) &
Verified code \& AI logs \\
\hline
Nguyen Van Tinh & 24125106 &
1B (Data split), 3A (Ridge), 3B (Lasso), 3C (Algebra Evidence) &
Verify mathematical problems \\
\hline
Nguyen Le Quoc Huy & 24125100 &
2B (Ridge), 2D (Model comparison), 2E (Stability Check), 4A (Paper research) &
Verify the project workflow \& code \\
\hline
\end{tabular}
\endgroup
```
# Abstract {-}

This study aims to predict pre-operative log prostate-specific antigen (`lpsa`) from clinical measurements, comparing models on prediction error, coefficient stability, and interpretability under multicollinearity. We evaluated OLS, Ridge, Lasso, and Elastic Net regressions using five-fold cross-validation on original linear predictors and fixed random ReLU features. Cross-validation initially favored Lasso (CV-MSE = 0.5790) for its sparse selection of five variables. However, holdout evaluation revealed that the OLS model achieved the lowest test RMSE (0.6713) and MAE (0.5736). Furthermore, original linear features consistently outperformed the nonlinear ReLU representations across all models. Despite the OLS superiority in pure holdout accuracy, we recommend the lasso-dominant elastic net ($\alpha = 0.9$) for clinical deployment, as it provides a parsimonious, highly interpretable model that isolates key biological drivers of PSA with only a modest trade-off in error.

<!-- list of model: -->
<!-- - OLS : Outperforms on test set , but dense and complex model (retains all 8 predictors).  -->
<!-- - Ridge (Original): Solves multicollinearity (reduces condition number from 20.6 to 8.9) and stabilizes coefficient signs, but lacks feature selection (still keeps 8 predictors) and has higher test error than OLS. -->
<!-- - Lasso / Elastic Net ($\alpha = 0.9$): Best training performance (lowest CV-MSE). Highly interpretable and parsimonious model (retains only 5 core biological predictors). -->
<!-- - Fixed ReLU Models: Worst holdout performance across all penalties. -->
# Reproducible setup {-}

```{r setup, warning=FALSE, message = FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr", "car", "ggplot2")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)
library(car)
library(ggplot2)
# `params` exists only while knitting; the fallback
# keeps interactive chunk runs working.
has_params <- exists("params") && !is.null(params$group_number)
group_number <- if (has_params) as.integer(params$group_number) else 4L
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```

## Shared helper functions {-}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
analysis_model <- function(x, y, alpha, foldid) {
  model <- fit_cv_glmnet(x = x, y = y, alpha=alpha, foldid = foldid)
  list(
    model = model,
    lambda_min = model$lambda.min,
    mse_min = model$cvm[model$lambda == model$lambda.min],
    cond_min = safe_condition_numbers(x, lambda = model$lambda.min)[2],
    coef_min = coef(model, "lambda.min"),
    coef_min_norm = sqrt(sum(coef(model, "lambda.min")[-1]^2)),
    nonzero_min = count_nonzero(model, s="lambda.min"),
    lambda_1se = model$lambda.1se,
    cond_1se = safe_condition_numbers(x, lambda = model$lambda.1se)[2],
    mse_1se = model$cvm[model$lambda == model$lambda.1se],
    coef_1se = coef(model, "lambda.1se"),
    coef_1se_norm = sqrt(sum(coef(model, "lambda.1se")[-1]^2)),
    nonzero_1se = count_nonzero(model, s="lambda.1se")
  )
}
```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration}

data <- read.table("data/airfoil_self_noise.dat", header = FALSE)
colnames(data) <- c("frequency", "attack_angle", "chord_length", "free_stream_velocity", "suction_side_displacement_thickness", "scaled_sound_pressure")

# Always (re)written from the .dat source so the cached CSV can never drift
# from the column-naming convention fixed above.
write.csv(
  data,
  "data/airfoil_self_noise.csv",
  row.names = FALSE
)

str(data)
config <- list(
  file = "airfoil_self_noise.csv",
  response = "scaled_sound_pressure",
  excluded = "scaled_sound_pressure"
)
data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```


```{r data-overview, echo=FALSE, warning=FALSE, message=FALSE}


ggplot(data, aes(x = chord_length)) +
  geom_histogram(bins = 30)



# data$log_frequency <- log(data$frequency)
# hist(data$frequency)
# hist(data$log_frequency)


data_overview <- tibble(
  Variable = names(data_raw),
  Role     = ifelse(names(data_raw) == config$response, "Response", "Predictor"),
  Type     = vapply(data_raw, function(x) class(x)[1L], character(1)),
  Distinct = vapply(data_raw, dplyr::n_distinct, integer(1), na.rm = TRUE),
  Missing  = vapply(data_raw, function(x) sum(is.na(x)), integer(1))
)

kable(
  data_overview,
  caption = paste("Response and candidate predictors, with the integrity checks",
                  "run before the split.")
)

n_duplicate_rows <- sum(duplicated(analysis_data))
n_missing_total  <- sum(is.na(analysis_data))
n_levels_freq    <- dplyr::n_distinct(data_raw$frequency)
n_levels_chord   <- dplyr::n_distinct(data_raw$chord_length)
n_levels_vel     <- dplyr::n_distinct(data_raw$free_stream_velocity)
n_levels_angle   <- dplyr::n_distinct(data_raw$attack_angle)
n_levels_thick   <- dplyr::n_distinct(data_raw$suction_side_displacement_thickness)

kable(
  data.frame(
    Check = c("Missing values (total cells)", "Fully duplicated rows"),
    Count = c(n_missing_total, n_duplicate_rows)
  ),
  caption = "Missing-value and duplicate-row checks on the raw data."
)
```

The dataset contains `r nrow(data_raw)` observations of `r ncol(data_raw)` numeric
variables: five predictors recorded during NACA 0012 airfoil wind-tunnel tests
and the response `scaled_sound_pressure`, the scaled sound pressure level in
decibels. No missing values or duplicated rows are present (table above), so no
imputation or de-duplication is required before the split.

The `Distinct` column in Table 1 shows that the five predictors are not equally
"continuous" in practice. `frequency`, `chord_length`, and `free_stream_velocity`
take only `r n_levels_freq`, `r n_levels_chord`, and `r n_levels_vel` distinct
values respectively - these are the fixed 1/3-octave frequency bands, airfoil
chord lengths, and wind-tunnel speeds used to design the experiment, so each
behaves more like an ordinal experimental factor than a freely varying
measurement. `attack_angle` (`r n_levels_angle` distinct values) and
`suction_side_displacement_thickness` (`r n_levels_thick` distinct values) are
closer to genuinely continuous quantities. This low cardinality on three of the
five predictors is expected to show up later as visibly clustered histograms and
boxplots (1C) rather than smooth continuous distributions, and is a property of
the experimental design rather than a data-quality issue.

`data/data_description.md` (the accompanying UCI variable sheet) lists
`attack_angle` as "Binary", which does not match what is observed here - the
variable takes `r n_levels_angle` distinct values between
`r round(min(data_raw$attack_angle), 1)` and `r round(max(data_raw$attack_angle), 1)`
degrees. This looks like a labelling error in the source sheet, and `attack_angle`
is treated as continuous throughout, consistent with the assignment brief's own
description of the dataset ("angle of attack" as a continuous input).



## 1B. Split and preprocessing plan

```{r split-and-scale}
split <- split_rows(nrow(analysis_data), seed = seeds$split)
train_id <- split$train
test_id <- split$test
train_df <- analysis_data[train_id, ]

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]

foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

To ensure a fair comparison among OLS, ridge, lasso, and elastic net, all preprocessing and model development steps were performed using only the training data. The holdout test set was reserved exclusively for the final evaluation and was not used during model fitting, model selection, or hyperparameter tuning.

### Data split

The observations were randomly partitioned into training and test sets using the custom `split_rows()` function with a training proportion of 80%. For Group `r sprintf("%02d", group_number)`, the split seed was **`r seeds$split`**, ensuring that the same partition can be reproduced in future analyses.

```{r echo=FALSE, warning=FALSE, message=FALSE}
kable(
  data.frame(
    Observation = c("Training", "Test observations"),
    Value = c(length(train_id), length(test_id))
  )
)

obs <- c(20,35,20,25); probs <- rep(1/4,4)
chisq.test(x = obs, p = probs)
```

The training data were used for exploratory analysis, preprocessing, model fitting, and cross-validation, while the holdout test data remained untouched until the final evaluation.

### Predictor standardization

Since ridge regression, lasso, and elastic net are sensitive to the scale of predictor variables, all predictors were standardized before model fitting. The centering and scaling parameters were estimated **only from the training data**, and these same values were subsequently applied to the test predictors.

Using training-derived scaling parameters prevents information from the holdout observations from influencing the preprocessing stage and ensures that the test set remains an unbiased evaluation dataset.

### Five-fold cross-validation

Model tuning was carried out using five-fold cross-validation on the training data only. Fold assignments were generated using the fixed cross-validation seed **`r seeds$cv`** and were shared across all regularized models. Reusing identical folds ensures that differences in cross-validation performance are attributable to the models themselves rather than random variation in fold assignment.

### Missing-value handling

The dataset was checked for missing values before model fitting.

```{r echo=FALSE, warning=FALSE, message=FALSE}
kable(data.frame(Total_missing_values = sum(is.na(analysis_data))))
```

No missing values were detected; therefore, no imputation was required. If missing values had been present, imputation statistics would have been estimated using only the training data and then applied unchanged to the test data.

### Invalid-value check

```{r invalid-values, echo=FALSE, warning=FALSE, message=FALSE}
implausible_checks <- data.frame(
  Variable   = c("frequency", "attack_angle", "chord_length",
                 "free_stream_velocity", "suction_side_displacement_thickness"),
  Rule       = c("> 0 Hz", ">= 0 deg", "> 0 m", "> 0 m/s", "> 0 m"),
  Violations = c(
    sum(train_df$frequency <= 0),
    sum(train_df$attack_angle < 0),
    sum(train_df$chord_length <= 0),
    sum(train_df$free_stream_velocity <= 0),
    sum(train_df$suction_side_displacement_thickness <= 0)
  )
)

kable(
  implausible_checks,
  caption = "Physical-plausibility checks on the training predictors."
)
```

Beyond the statistical IQR screen below, each predictor was also checked against
the range it is physically allowed to take: a frequency, chord length, velocity,
or boundary-layer thickness cannot be zero or negative, and an angle of attack
cannot be negative. No violations are found on the training set, which is
expected for an instrument-recorded wind-tunnel experiment rather than
manually-entered data.

### Outlier detection and handling

```{r outlier-detection, echo=FALSE, warning=FALSE, message=FALSE}
detect_outliers_iqr <- function(x) {
  q     <- stats::quantile(x, probs = c(0.25, 0.75), na.rm = TRUE)
  fence <- 1.5 * (q[2] - q[1])
  which(x < q[1] - fence | x > q[2] + fence)
}

outlier_cols   <- c(predictors, config$response)
outlier_counts <- vapply(
  train_df[, outlier_cols],
  function(col) length(detect_outliers_iqr(col)),
  integer(1)
)

outlier_table <- data.frame(
  Variable  = outlier_cols,
  Outliers  = as.integer(outlier_counts),
  Percent   = round(100 * unname(outlier_counts) / nrow(train_df), 1),
  row.names = NULL
)

kable(
  outlier_table,
  caption = paste("Training-set values flagged by the 1.5×IQR rule,",
                   "computed on the training split only.")
)

freq_outliers  <- outlier_counts[["frequency"]]
thick_outliers <- outlier_counts[["suction_side_displacement_thickness"]]
angle_outliers <- outlier_counts[["attack_angle"]]
resp_outliers  <- outlier_counts[[config$response]]
```

Outliers were screened with the standard $1.5\times\text{IQR}$ fence, applied to
each variable using **training-set quantiles only** so that no information from
the held-out test rows enters the check. `chord_length` and
`free_stream_velocity` contribute zero flagged points, because - as noted in
1A - each takes only a handful of fixed values, so every observation sits well
inside its own IQR fence by construction. `frequency`, `attack_angle`, and
`suction_side_displacement_thickness` each contribute a nontrivial share
(`r freq_outliers`, `r angle_outliers`, and `r thick_outliers` training rows
respectively), and `r resp_outliers` response values are flagged.

These flagged points are **retained rather than removed**. The airfoil data
come from a controlled wind-tunnel experiment, not a survey or manual data-entry
process, so there is no mechanism here for typographical or transcription
errors; every extreme value is a physically valid test condition (a high
frequency band, a large angle of attack, a thin boundary layer) rather than a
data-quality defect. The IQR rule is doing exactly what it is designed to do on
a right-skewed engineering variable: flagging the upper tail of a real physical
distribution, not finding mistakes. Dropping these rows would bias the training
sample away from precisely the loud, high-frequency regime the model is meant to
predict, so all `r nrow(train_df)` training rows are kept for model fitting in
Problem 2.

### Transformation check

```{r transform-check, echo=FALSE, warning=FALSE, message=FALSE}
skewness <- function(x) {
  z <- x - mean(x)
  mean(z^3) / (mean(z^2))^1.5
}

skew_raw <- vapply(train_df[, predictors], skewness, numeric(1))
skew_log <- c(
  frequency = skewness(log(train_df$frequency)),
  suction_side_displacement_thickness =
    skewness(log(train_df$suction_side_displacement_thickness))
)

skew_table <- data.frame(
  Variable           = names(skew_raw),
  Skewness           = unname(round(skew_raw, 3)),
  Skewness_after_log = c(
    round(skew_log[["frequency"]], 3), NA, NA,
    round(skew_log[["suction_side_displacement_thickness"]], 3), NA
  ),
  row.names = NULL
)

kable(
  skew_table,
  caption = paste("Training-set skewness of each predictor, and after a log",
                   "transform for the two most skewed variables.")
)

# Does log-transforming the two skewed predictors improve a plain linear fit?
ols_raw_form <- reformulate(predictors, response = config$response)
ols_raw_fit  <- lm(ols_raw_form, data = train_df)

train_df_log <- train_df
train_df_log$frequency <- log(train_df_log$frequency)
train_df_log$suction_side_displacement_thickness <-
  log(train_df_log$suction_side_displacement_thickness)
ols_log_fit <- lm(ols_raw_form, data = train_df_log)

r2_raw <- summary(ols_raw_fit)$r.squared
r2_log <- summary(ols_log_fit)$r.squared

# Is the frequency effect itself curved rather than log-shaped?
r2_freq_linear <- summary(lm(train_df[[config$response]] ~ train_df$frequency))$r.squared
r2_freq_quad   <- summary(lm(train_df[[config$response]] ~ train_df$frequency +
                                I(train_df$frequency^2)))$r.squared
```

`frequency` and `suction_side_displacement_thickness` are the two right-skewed
predictors identified above (skewness `r round(skew_raw[["frequency"]], 2)` and
`r round(skew_raw[["suction_side_displacement_thickness"]], 2)`). A log
transform brings both close to symmetric on the training set (skewness
`r round(skew_log[["frequency"]], 2)` and
`r round(skew_log[["suction_side_displacement_thickness"]], 2)`), which is the
usual motivation for log-transforming a positive, right-skewed engineering
measurement. `attack_angle` is also mildly right-skewed (skewness
`r round(skew_raw[["attack_angle"]], 2)`, Table above) but far less than
`suction_side_displacement_thickness`, so it is not treated as a transform
candidate here.

However, symmetrizing the marginal distribution is not the same as improving
the fit, and it is checked rather than assumed. Refitting a plain linear model
on the training data with these two predictors log-transformed gives
$R^2 = `r round(r2_log, 4)`$, against $R^2 = `r round(r2_raw, 4)`$ for the
untransformed predictors - the log transform does not improve, and mildly
*hurts*, the linear fit. A separate check on `frequency` alone points to why: a
quadratic term explains more of the response than a linear term
($R^2 = `r round(r2_freq_quad, 4)`$ vs. $R^2 = `r round(r2_freq_linear, 4)`$),
so the frequency-response relationship is curved in a way a log transform does
not match - consistent with the assignment brief's own framing of this dataset
as having "few features but nonlinear relationships."

Given this evidence, no log transform is applied. All predictors are carried
forward on their original scale into Problem 2, with only the training-derived
standardization described above; the nonlinearity documented here is noted as a
limitation of the linear and regularized-linear models used there, rather than
something a simple transform resolves.

### Data cleaning summary

```{r cleaning-summary, echo=FALSE, warning=FALSE, message=FALSE}
cleaning_summary <- data.frame(
  Check = c("Missing values", "Duplicate rows", "Invalid values",
            "IQR outliers", "Skewness", "Categorical variables",
            "Predictor scaling"),
  Result = c(
    paste(n_missing_total, "cells"),
    paste(n_duplicate_rows, "rows"),
    paste(sum(implausible_checks$Violations), "violations"),
    paste(sum(outlier_counts), "variable-level flags (a row may be flagged on more than one variable)"),
    "frequency and suction_side_displacement_thickness right-skewed",
    "none present (all 5 predictors numeric)",
    "required for ridge / lasso / elastic net"
  ),
  Decision = c(
    "No imputation needed",
    "No de-duplication needed",
    "No rows dropped",
    "Retained (valid experimental conditions, not errors)",
    "No log transform applied (does not improve the linear fit)",
    "No encoding needed",
    "Applied in Problem 2, fit on training data only"
  )
)

kable(
  cleaning_summary,
  caption = "Summary of data-cleaning checks and decisions carried into Problem 2."
)
```

### Leakage control

Several safeguards were implemented to prevent information leakage throughout the analysis:

- The train–test split was performed before any preprocessing.
- Predictor standardization used statistics estimated from the training data only.
- The invalid-value, outlier, and transformation checks above used training-set values, quantiles, and model fits only.
- Five-fold cross-validation was conducted exclusively within the training set.
- Hyperparameter tuning for ridge, lasso, and elastic net relied solely on training cross-validation error.
- The holdout responses (`y_test`) were not accessed until the final evaluation stage.

As a result, the holdout test set provides an unbiased assessment of predictive performance because no information from the test observations entered the model construction or tuning process.


## 1C. Data exploration

### Pairwise correlation among predictors
```{r eda-cor, echo=FALSE, warning=FALSE, message=FALSE}
# Pairwise correlation among predictors
cor_matrix <- cor(
  train_df[, predictors],
  use = "complete.obs"
)

knitr::kable(
  round(cor_matrix, 2),
  caption = "Pairwise correlation matrix among predictors"
)
```

```{r eda-heatmap, echo=FALSE, warning=FALSE, message=FALSE}
# Correlation heatmap
layout(matrix(c(1, 2), nrow = 1), widths = c(8.5, 1.5))
par(mar = c(5, 5, 2, 1))
cor_matrix <- cor(train_df[, predictors], use = "complete.obs")

image(1:ncol(cor_matrix), 1:ncol(cor_matrix), cor_matrix, 
      axes = FALSE, xlab = "", ylab = "", 
      col = heat.colors(1000, rev=TRUE), main = "Correlation Heatmap of Predictors")

axis(1, at = 1:ncol(cor_matrix), labels = colnames(cor_matrix), las = 2)
axis(2, at = 1:ncol(cor_matrix), labels = colnames(cor_matrix), las = 1)

par(mar = c(5, 1, 2, 3)) 
legend_scale <- matrix(seq(0, 1, length.out = 100), nrow = 1)
image(1, seq(0, 1, length.out = 100), legend_scale, 
      col = heat.colors(20, rev=TRUE), axes = FALSE, xlab = "", ylab = "")

axis(4, at = seq(0, 1, by = 0.1), las = 1, cex.axis = 0.8)
```

```{r eda-cor-summary, echo=FALSE, warning=FALSE, message=FALSE}
cor_off_diag <- cor_matrix
diag(cor_off_diag) <- NA
max_pair_idx <- which(abs(cor_off_diag) == max(abs(cor_off_diag), na.rm = TRUE), arr.ind = TRUE)[1, ]
max_pair_vars <- rownames(cor_matrix)[max_pair_idx]
max_pair_val  <- cor_matrix[max_pair_idx[1], max_pair_idx[2]]
```

Predictor-predictor correlations are weak across most pairs. The one exception
is `r max_pair_vars[1]` and `r max_pair_vars[2]`, correlated at
$r \approx `r round(max_pair_val, 2)`$ - a moderate relationship, plausible
given both describe the airfoil's boundary-layer geometry at the trailing edge.
No other pair exceeds $|r| \approx 0.5$. This single moderate correlation is
revisited with the VIF check below and is the main multicollinearity signal
this dataset offers ridge and lasso a reason to address in Problem 2.

### Response Distribution

```{r eda-response-dist, echo=FALSE, warning=FALSE, message=FALSE}

hist(y_train)
summary <- as.matrix(summary(y_train))

# ggplot(data, aes(x = frequency, y = scaled_sound_pressure)) +
#   geom_point()
kable(
  data.frame(
    Row = rownames(summary),
    Value = as.numeric(summary[,1])
  )
)

resp_skew <- skewness(y_train)
```

The response ranges from `r round(min(y_train), 1)` to `r round(max(y_train), 1)`
dB with a median of `r round(median(y_train), 1)` dB and mean
`r round(mean(y_train), 1)` dB. The histogram and the small gap between mean and
median show a mild left skew (skewness $\approx `r round(resp_skew, 2)`$): most
measurements cluster in the `r round(quantile(y_train, 0.25), 0)`-
`r round(quantile(y_train, 0.75), 0)` dB range, with a longer tail toward the
quieter (lower-dB) end rather than the loud end. The skew is mild enough that no
response transformation is applied - RMSE and MAE on the original dB scale
remain directly interpretable as sound-pressure prediction errors, which matters
for Problem 1's evaluation step.

### Predictor Histogram  & Boxplot

1. Histogram

```{r eda-hist, echo=FALSE, warning=FALSE, message=FALSE}
## Histogram
par(mfrow = c(2,4))

for (var in predictors) {
  hist(
    train_df[[var]],
    main = paste("Distribution of", var),
    xlab = var,
    breaks = 15
  )
}

par(mfrow = c(1,1))
```


2. Boxplot

```{r eda-boxplot, echo=FALSE, warning=FALSE, message=FALSE}
## Boxplot
par(mfrow = c(2,4))
for (var in predictors) {
  boxplot(
    train_df[[var]],
    main = var
  )
}
par(mfrow = c(1,1))


```

The histograms and boxplots make the 1A distinct-value counts visible. `chord_length`
and `free_stream_velocity` appear as `r n_levels_chord` and `r n_levels_vel`
narrow, evenly spaced spikes rather than a smooth spread, and `frequency` shows
`r n_levels_freq` spikes concentrated at the low end with a long right tail -
these are the fixed experimental levels described in 1A, not gaps in the data.
`attack_angle` and `suction_side_displacement_thickness` are closer to
continuous, right-skewed shapes, matching the skewness values quantified in 1B.
The boxplots' upper whiskers and flagged points for these two variables are the
same training-set values already counted in the 1B outlier check, retained for
the same reason: they are valid high-frequency / high-angle test conditions, not
data errors.

### Correlation between predictors and response

```{r eda-cor-res-pred, echo=FALSE, warning=FALSE, message=FALSE}
cor_response <- cor(
  train_df[, predictors],
  train_df[[config$response]]
)
colnames(cor_response) <- config$response

knitr::kable(
  round(cor_response, 2),
  caption = "Correlation between predictors and scaled_sound_pressure"
)

strongest_pred <- rownames(cor_response)[which.max(abs(cor_response[, 1]))]
strongest_cor  <- cor_response[strongest_pred, 1]
```

Every predictor's marginal correlation with the response is weak in magnitude
(all $|r| < 0.4$). `r strongest_pred` is the strongest single linear predictor
($r \approx `r round(strongest_cor, 2)`$), followed by
`suction_side_displacement_thickness`; `free_stream_velocity` is weakest and the
only predictor positively correlated with the response. No individual predictor
comes close to explaining the response on its own, which is consistent with the
assignment brief's description of this dataset as having "few features but
nonlinear relationships" - a linear combination of all five predictors, rather
than any single one, is needed, and Problem 2's regularized models are expected
to combine these weak individual signals rather than select one dominant
predictor and discard the rest.

###  Check multicollinearity using VIF
```{r eda-vif, echo=FALSE, warning=FALSE, message=FALSE}

response <- config$response

vif_model <- lm(
  as.formula(
    paste0("`", response, "` ~ .")
  ),
  data = train_df[, c(response, predictors)]
)

vif_values <- vif(vif_model)

knitr::kable(
  data.frame(VIF = round(vif_values, 2)),
  caption = "Variance Inflation Factor (VIF)"
)

max_vif_1c      <- max(vif_values)
max_vif_var_1c  <- names(which.max(vif_values))
```

Every VIF is comfortably below the conventional threshold of 5, with
`r max_vif_var_1c` the highest at `r round(max_vif_1c, 2)`. This confirms the
correlation-matrix finding above: the airfoil predictors share only mild,
tolerable overlap, so no predictor needs to be dropped for redundancy, and OLS
in Problem 2 is not expected to suffer severe coefficient instability from
multicollinearity. The moderate `attack_angle` / `suction_side_displacement_thickness`
correlation is still the largest source of shared variance in the data and
remains worth tracking through ridge and lasso, but it falls well short of the
kind of collinearity (VIF $\gg$ 5) that would make variable selection a
necessity rather than a convenience.

###  Predictor-response plots
```{r eda-pre-res-plot, echo=FALSE, warning=FALSE, message=FALSE}

par(mfrow = c(2,4))

for (var in predictors) {
  plot(
    train_df[[var]],
    train_df[[config$response]],
    xlab = var,
    ylab = config$response,
    main = paste(var, "vs", config$response),
    pch = 19
  )

  abline(
    lm(train_df[[config$response]] ~ train_df[[var]]),
    col = "red"
  )
}

par(mfrow = c(1,1))
```

The fitted red lines are visibly weak for every panel - the same
$|r| < 0.4$ ceiling seen in the correlation table above - but the point clouds
are not simply noise around those lines. The `frequency` panel in particular
shows the response bending down over its range rather than following the
straight line, echoing the linear-vs-quadratic $R^2$ comparison from 1B: the
straight-line fit misses curvature that a richer model could capture. The
`chord_length` and `free_stream_velocity` panels show their points arranged in
`r n_levels_chord` and `r n_levels_vel` vertical strips (their fixed
experimental levels), each strip itself spanning a wide range of the response -
visual confirmation that these two variables alone say little about the
outcome, and any predictive power must come from combining all five predictors
together, which is exactly what Problem 2's models are built to do.

### Question

Taken together, the checks in 1A-1C describe a clean but weak-signal regression
problem, and that combination shapes what the rest of Problem 1 needs to
prioritise.

**Data quality is not the challenge here.** There are no missing values, no
duplicated rows, and the "outliers" flagged by the 1.5×IQR rule are legitimate
wind-tunnel test conditions from a controlled experiment, not data-entry errors
- so cleaning required no row deletions or imputation, only the decision (1B) to
retain every training row and to skip a log transform after checking that it
does not help the linear fit.

**The predictors are almost uncorrelated with each other.** All VIFs sit well
under 5, and the one moderately correlated pair
(`attack_angle` / `suction_side_displacement_thickness`, $r \approx `r round(max_pair_val, 2)`$)
falls short of severe multicollinearity. This means ridge and lasso in Problem 2
are not being asked to rescue an ill-conditioned design matrix the way they
would with, say, several near-duplicate predictors; their main job here is
tuning the bias-variance trade-off and (for lasso) testing whether any predictor
can be dropped outright, rather than resolving instability.

**The predictors are individually weak but the relationships are nonlinear.**
Every predictor-response correlation is under 0.4 in magnitude, and no single
predictor separates high- from low-noise conditions on its own (the
`chord_length` and `free_stream_velocity` panels above make this explicit, with
each fixed level covering a wide spread of the response). At the same time,
`frequency` fits a quadratic term noticeably better than a straight line
(1B), and three of the five predictors are effectively small sets of
experimental design levels (1A) rather than smooth continuous measurements.
Both point to the same conclusion the assignment brief anticipates for this
dataset: the frequency-response relationship, and likely others, are genuinely
nonlinear, so a **linear baseline (OLS) is expected to leave a meaningful
amount of predictable structure on the table**, and its residuals in Problem 2
should be inspected for exactly this kind of unmodelled curvature rather than
assumed to be pure noise.
