---
title: "ASESII 24A02 Final Project - Part 2"
subtitle: "2-way ANOVA method"
author: |
  | Group 02
  | Nguyen Thi Yen Nhi (24125071)
  | Nguyen Khang Thinh (24125104)
  | Nguyen Van Tinh (24125106)
  | Nguyen Le Quoc Huy (24125100)
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
link-citations: true
nocite: |
  @loshchilov2019decoupledweightdecayregularization, @krogh1992simple
params:
  group_number: 02
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere}
  - \usepackage{float}
  - \usepackage{longtable}
--- 

# Abstract

```{r setup, echo=FALSE, warning=FALSE, message = FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr", "car", "ggplot2", "dplyr", "kableExtra")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)
library(car)
library(ggplot2)
library(kableExtra)
library(dplyr)
# `params` exists only while knitting; the fallback
# keeps interactive chunk runs working.
has_params <- exists("params") && !is.null(params$group_number)
group_number <- if (has_params) as.integer(params$group_number) else 4L
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```


```{r helpers, echo=FALSE}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
analysis_model <- function(x, y, alpha, foldid) {
  model <- fit_cv_glmnet(x = x, y = y, alpha=alpha, foldid = foldid)
  list(
    model = model,
    lambda_min = model$lambda.min,
    mse_min = model$cvm[model$lambda == model$lambda.min],
    cond_min = safe_condition_numbers(x, lambda = model$lambda.min)[2],
    coef_min = coef(model, "lambda.min"),
    coef_min_norm = sqrt(sum(coef(model, "lambda.min")[-1]^2)),
    nonzero_min = count_nonzero(model, s="lambda.min"),
    lambda_1se = model$lambda.1se,
    cond_1se = safe_condition_numbers(x, lambda = model$lambda.1se)[2],
    mse_1se = model$cvm[model$lambda == model$lambda.1se],
    coef_1se = coef(model, "lambda.1se"),
    coef_1se_norm = sqrt(sum(coef(model, "lambda.1se")[-1]^2)),
    nonzero_1se = count_nonzero(model, s="lambda.1se")
  )
}
```

# Dataset selection

## Data Overview

```{r data-configuration}

data <- read.csv("data/penguins_size.csv", header = TRUE)

# Treat "." as a missing value
data[data == "."] <- NA

config <- list(
  file = "penguins_size.csv",
  response = "body_mass_g",
  factors = c("species", "sex")
)

data_raw <- data

analysis_data <- data_raw[
  ,
  c(config$response, config$factors),
  drop = FALSE
]

colSums(is.na(analysis_data))

analysis_data <- na.omit(analysis_data)

if (anyNA(analysis_data)) {
  stop("Missing values remain in the analysis data.")
}
```

```{r data-overview, echo=FALSE, warning=FALSE, message=FALSE}
#str(analysis_data)

#table(analysis_data$species)
#table(analysis_data$sex)

#table(
#  analysis_data$species,
#  analysis_data$sex
#)

data_overview <- tibble(
  Variable = names(data_raw),
  Role = ifelse(
    names(data_raw) == config$response,
    "Response",
    "Predictor"
  ),
  Type = vapply(
    data_raw,
    function(x) class(x)[1L],
    character(1)
  ),
  Distinct = vapply(
    data_raw,
    dplyr::n_distinct,
    integer(1),
    na.rm = TRUE
  ),
  Missing = vapply(
    data_raw,
    function(x) sum(is.na(x)),
    integer(1)
  )
)

kable(
  data_overview,
  caption = "Data Overview Before Missing-Value Removal"
)
# n_duplicate_rows <- sum(duplicated(analysis_data))

# n_duplicate_rows
```
For 2-Way ANOVA, this method requires the dataset to have at least 2 categorical variables along with a continuous response variable for the analysis.

In general, from our chosen dataset, we decide to evaluate the factors that are associated with penguin body mass. This dataset contains information about 344 penguins, including species, island, physical measurements such as culmen length, culmen depth, and flipper length, body mass, and sex. It is suitable for exploratory data analysis and 2-Way ANOVA analysis based on the biological and physical characteristics of penguins.

Specifically, the dataset contains 3 categorical variables: `species` (Adelie/Chinstrap/Gentoo), `island` (Torgersen/Biscoe/Dream), and `sex` (MALE/FEMALE). In this method, we select `species` and `sex` as the two categorical factors and `body_mass_g` as the continuous response variable. The analysis is used to test whether the two factors are associated with differences in mean body mass, as well as whether there is an interaction between the two factors.
  
## Research Questions & Hypotheses

We investigate whether penguin species and sex are associated with differences in mean body mass, and whether there is an interaction between the two factors.

We follow the 2-way ANOVA model:


$$
Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}
$$

with $\epsilon_{ijk} \sim N(0,\sigma^2)$.

- $\mu$ is the overall mean.
- $\alpha_i$ is the effect of the $i$-th level of Factor A, `species`.
- $\beta_j$ is the effect of the $j$-th level of Factor B, `sex`.
- $(\alpha\beta)_{ij}$ is the interaction effect between Factor A and Factor B.
- $\epsilon_{ijk}$ is the random error term.

### Factor A: Penguin Species

**Research Question:**  
Is penguin species associated with differences in mean body mass?

\[
\begin{cases}
H_{0a}: \alpha_1 = \alpha_2 =  \alpha_3 = 0\\
H_{1a}: \text{Not all } \alpha_i \text{ are equal to }0.
\end{cases}
\]

### Factor B: Penguin Sex

**Research Question:**  
Is the sex of penguin associated with differences in mean body mass?

\[
\begin{cases}
H_{0b}: \beta_1 = \beta_2 = 0\\
H_{1b}: \text{Not all } \beta_j \text{ are equal to } 0.
\end{cases}
\]

### Interaction Effect

**Research Question:**  
Does the association between sex and penguin body mass depend on the penguin species?

\[
\begin{cases}
H_{0ab}: (\alpha\beta)_{ij} = 0 \quad \text{for all } i,j\\
H_{1ab}: \text{Not all } (\alpha\beta)_{ij} \text{ are equal to } 0.
\end{cases}
\]

## Descriptive Statistics

### Data Summary Statistics
```{r data-response-summary, echo=FALSE}

# Overall summary statistics on response
overall_summary <- analysis_data %>%
  summarise(
    N = n(),
    Mean = mean(body_mass_g, na.rm = TRUE),
    Median = median(body_mass_g, na.rm = TRUE),
    Variance = var(body_mass_g, na.rm = TRUE),
    SD = sd(body_mass_g, na.rm = TRUE),
    Min = min(body_mass_g, na.rm = TRUE),
    Max = max(body_mass_g, na.rm = TRUE)
  )

overall_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    caption = "Overall Summary Statistics of Penguin Body Mass",
    align = "c"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    position = "center"
  )

```

The dataset contains 333 observations of penguin body mass after removing observations with missing values. The mean body mass is 4207.06 g, with a median of 4050 g, indicating that the body masses are centered around approximately 4.0 to 4.2 kg. The standard deviation is 805.22 g, showing moderate variability in body mass, while the observed values range from 2700 g to 6300 g. The mean being slightly higher than the median suggests a mild right skew in the overall response distribution.

```{r data-treatment-combination-summary, echo=FALSE}

# Summary statistics by treatment combination
group_summary <- analysis_data %>%
  group_by(
    species,
    sex
  ) %>%
  summarise(
    N = n(),
    Mean = mean(body_mass_g, na.rm = TRUE),
    Median = median(body_mass_g, na.rm = TRUE),
    Variance = var(body_mass_g, na.rm = TRUE),
    SD = sd(body_mass_g, na.rm = TRUE),
    Min = min(body_mass_g, na.rm = TRUE),
    Max = max(body_mass_g, na.rm = TRUE),
    .groups = "drop"
  )

group_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    caption = "Summary Statistics by Treatment Combination",
    col.names = c(
      "Species",
      "Sex",
      "N",
      "Mean",
      "Median",
      "Variance",
      "SD",
      "Min",
      "Max"
    ),
    align = "c"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    position = "center"
  )
```
The dataset contains three levels of species and two levels of sex. Male penguins generally have higher mean body mass than female penguins across all three species.

The difference in mean body mass between males and females varies across species. The difference is approximately 674.65 g for Adelie penguins, 411.76 g for Chinstrap penguins, and 805.10 g for Gentoo penguins.
This variation provides descriptive evidence of a potential interaction between species and sex.

Among the six treatment combinations, the Gentoo male group has the highest mean body mass (5484.84 g), while the Adelie female group has the lowest mean body mass (3368.84 g). The group variances range from
72,565.64 to 120,278.25, with the Adelie male group having the largest variance. Therefore, the homogeneity of variance assumption should be checked before conducting the two-way ANOVA.

### Visualization

#### Histogram of Penguin Body Mass

```{r data-hist, echo=FALSE, fig.width=6, fig.height=4, out.width="90%", fig.align="center"}
ggplot(analysis_data, aes(x = body_mass_g)) +
  geom_histogram(
    bins = 20,
    color = "black"
  ) +
  labs(
    title = "Distribution of Penguin Body Mass",
    x = "Body Mass",
    y = "Frequency"
  ) +
  theme_minimal()

body_mass_g <- data$body_mass_g

```
The histogram shows that penguin body masses range from approximately 2700 g to 6300 g, with most observations concentrated between 3000 g and 5000 g. The distribution shows a mild right skew, with a longer tail
toward higher body masses. This is consistent with the mean body mass (4207.06 g) being higher than the median (4050 g). The distribution also shows relatively high frequencies around 3500 to 4000 g, while fewer observations are found at the upper end of the body mass range.

#### Box Plot 

```{r data-boxplot, echo=FALSE, fig.width=6.5, fig.height=4.5, out.width="95%", fig.align="center"}

ggplot(
  analysis_data,
  aes(
    x = interaction(
      species,
      sex,
      sep = " x "
    ),
    y = body_mass_g
  )
) +
  geom_boxplot() +
  scale_x_discrete(
    labels = function(x) gsub(" x ", "\n", x)
  ) +
  labs(
    title = "Penguin Body Mass by Treatment Combination",
    x = "Treatment Combination",
    y = "Body Mass (g)"
  ) +
  theme_minimal()
```

The box plot compares the distribution of penguin body mass across the six treatment combinations of species and sex. Male penguins generally have higher median and mean body mass than female penguins within each species.

The difference between male and female body mass appears across all three species. The largest difference is observed for Adelie penguins, with mean body mass increasing from 3368.84 g for females to 4043.49 g for males. For Chinstrap penguins, the corresponding means are 3527.21 g and 3938.97 g, while for Gentoo penguins they are 4679.74 g and 5484.84 g.

The spread of the groups is not completely uniform, with some groups showing wider variability than others. A few potential low outliers are also visible, particularly in the Chinstrap female group. Overall, the
plot suggests that both species and sex may be associated with penguin body mass, while the differences in the male-female effect across species provide descriptive evidence of a possible interaction between the two factors.

#### Interaction Plot

```{r interaction-plot, echo=FALSE, fig.width=6, fig.height=4, out.width="90%", fig.align="center"}

ggplot(
  analysis_data,
  aes(
    x = sex,
    y = body_mass_g,
    group = species,
    linetype = species
  )
) +
  stat_summary(
    fun = mean,
    geom = "line"
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 3
  ) +
  labs(
    title = "Interaction Between Species and Sex",
    x = "Sex",
    y = "Mean Body Mass (g)",
    linetype = "Species"
  ) +
  theme_minimal()
```

The interaction plot shows the mean body mass for each combination of species and sex. Across all three species, male penguins have higher mean body mass than female penguins. The mean body mass increases by
approximately 674.66 g for Adelie, 411.76 g for Chinstrap, and 805.10 g for Gentoo penguins.

The lines are not perfectly parallel. In particular, the Adelie and Chinstrap lines cross between the female and male groups, while the Adelie and Gentoo lines show relatively similar slopes. This provides
some descriptive evidence that the difference in body mass between males and females may vary across species.

However, the interaction pattern appears relatively modest overall, and the plot alone cannot determine whether the interaction is statistically significant. The significance of the species-by-sex interaction will be formally assessed using the two-way ANOVA.

### Assumptions Check

#### Normality of Residuals

```{r fit-model, echo=FALSE}
# fit model for checking assumptions
model <- aov(
  body_mass_g ~
    species * sex,
  data = data
)

residual <- residuals(model)
```


```{r normality, fig.width=6, fig.height=4, out.width="90%", fig.align="center"}
qqnorm(residual)
qqline(residual)
```

```{r shapiro}
shapiro.test(residual)
```

The normality assumption was assessed using the normal Q-Q plot and the Shapiro-Wilk test. The hypotheses for
the Shapiro-Wilk test are:


\[
\begin{cases}
H_0: \text{The residuals follow a normal distribution}\\
H_1: \text{The residuals do not follow a normal distribution}
\end{cases}
\]


The Shapiro-Wilk test resulted in

$$
W = 0.99776,\qquad p = 0.9367.
$$

Since the p-value is greater than 0.05, we fail to reject $H_0$. Therefore, there is no statistical evidence that the ANOVA residuals deviate from a normal distribution.

The Q-Q plot also shows that the residuals generally follow the reference line, with only minor deviations at the tails. The Shapiro-Wilk test gives $W=0.99776$ with a p-value of $0.9367$, providing further support for the normality assumption.

Overall, the residuals show no substantial departure from normality, and the normality assumption is considered satisfied for the subsequent two-way ANOVA.

#### Homogeneity of Variances 

```{r homogeneity}
leveneTest(
  body_mass_g ~
    species * sex,
  data = analysis_data
)
```

The homogeneity of variance assumption was assessed using the treatment group box plots and Levene's test. The six treatment combinations are formed by `species` and `sex`.

The hypotheses are:


\[
\begin{cases}
H_0:
\sigma_1^2 = \sigma_2^2 = \sigma_3^2 = \cdots =\sigma_6^2\\
H_1:
\text{At least one treatment group has a different variance}
\end{cases}
\]

Levene's test produced

$$
F(5,327) = 1.3908,\qquad p = 0.2272
$$
Since the p-value is greater than 0.05, we fail to reject $H_0$. Therefore, there is no statistical evidence that the variances differ across the six treatment combinations.

The group variances range from approximately 72,565.64 to 120,278.25, with the Adelie male group having the largest variance and the Adelie female group having the smallest variance. Although the group variances
are not identical, Levene's test suggests that the differences are not statistically significant. Therefore, the homogeneity of variance assumption is considered satisfied for the subsequent two-way ANOVA.

#### Independence of Errors

```{r independence-check}
sum(duplicated(analysis_data))
```

The independence assumption requires that the error terms are independent:

$$
\epsilon_{ijk} \overset{ind}{\sim} N(0,\sigma^2).
$$

The independence assumption was assessed based on the structure of the dataset. Each observation represents an individual penguin, and the dataset does not contain repeated measurements or a time-dependent
structure. Although some observations have identical values for the selected variables (`body_mass_g`, `species`, and `sex`), these do not necessarily represent duplicate penguins, since different penguins can
naturally have the same body mass and characteristics.

Therefore, independence is considered reasonable based on the available dataset information. However, it cannot be formally verified because the dataset does not provide a unique identifier or detailed information about the sampling procedure.

### Overall Assessment

The independence assumption is considered reasonable based on the structure of the dataset. Each observation represents an individual penguin, and there is no time-dependent or repeated-measurement structure identified in the analysis data. However, independence cannot be formally verified because the dataset does not provide a unique identifier or detailed information about the sampling procedure.

The normality assumption is considered satisfied based on both the Q-Q plot and the Shapiro-Wilk test ($p=0.9367$), which provides no statistical evidence of a departure from normality. The homogeneity of variance assumption is also considered satisfied based on the non-significant Levene's test ($p=0.2272$).

Overall, the assumptions are considered reasonably adequate for conducting the two-way ANOVA. No substantial concerns are observed regarding normality or homogeneity of variance, while the independence assumption is considered reasonable based on the available dataset information.
